---
title: "forest-snow-fire-manuscript"
output: bookdown::word_document2
date: "2023-03-20"
geometry: margin=1.0in
font-size: 12pt
editor_options: 
  markdown: 
    wrap: 100
bibliography: references.bib
csl: philtransrscb.csl
header-includes:
    - \usepackage{setspace}\doublespacing
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/jgoldman/Work/PhD/forest-snow-fire-interactions/")

library(tidyverse)
library(kableExtra)
library(bookdown)
```

# Abstract

**The one question is**: Climate warming is expected to increase fire severity, while reducing
snowfall, and causing earlier snowmelt and lengthening summer drought and fire season duration. To
date, the relationship between snow cover dynamics and wildfire severity remains uncertain.

**Here we do**: We used structural equation modelling to test a hypothesized causal structure
between forest structure, snow cover dynamics, and wildfire burn severity for 689 fires that burn in
Ontario's boreal forest between 2002-2019. 

**What we found**: We found that the relationship between stand structure and burn severity was
strongly mediated by snow cover, and that the strength of the relationships differed across spatial
scales.

**How it matters**:

# Introduction

Earlier snowmelt, decreased snowpack, and earlier spring moisture amounts are associated with longer
warm season droughts, with implications for the moisture availability in both spring and summer
[@gergel.etal2017]. Similar to soils, litter fuels (i.e., surface fuels) located on the forest floor
tend to recharge (i.e., water added to soil/litter moisture storage) during the cool season and dry
out following snowmelt [@estes2012]. This process is facilitated by the amount of water contained in
the snowpack. Snowpack allows snow-dominated temperate ecosystems to store more water than what
could be contained in the soil alone. Therefore, snow cover may determine the extent to which fine
fuels can recharge. Additionally, the release of water from the snowpack coincides with the spring
growing season and ensures moist condition in spring, as peak moisture is synchronous with snow
disappearance date [@harpoldSoilMoistureResponse2015]. However, snowmelt is heterogeneous at the
landscape scale, and the timing of soil drying (and potentially litter fuels) following snowmelt can
vary by several weeks [@harpoldSoilMoistureResponse2015, @hopmans.etal_2012] resulting in landscape
patches of different moisture levels.

Heterogeneity in snowpack amount and snowmelt are directly influenced by forest structural diversity
-- the three-dimensional variation in the forest (i.e., age, canopy height, biomass). Forest
structure modulates snow accumulation and disappearance by increasing canopy interception, snow
ablation, and sublimation [@schneiderTreeSpatialPatterns2019, @rothForestImpactsSnow2017,
@hojatimalekshah.etal2023]. In dense forests, such as the balsam fir and black spruce stands
characteristic of the eastern boreal forest, snow intercepted by the canopy decreases snowpack
[@balland2006] Further, relatively lower rates of solar radiation penetrating the canopy reduces
snow ablation [@varholaForestCanopyEffects2010]. In less dense forests, such as the jack pine stands
found in the central boreal forest, sparse canopy cover and increased distance between trees leads
to more snowpack [@balland2006]. However, solar radiation is higher in more open canopies resulting
in higher rates of ablation. Changes in the snowpack energy balance due to forest structure directly
alters spring snowmelt timing [@lundquistLowerForestDensity2013
@dickerson-langeSnowDisappearanceTiming2017 @musselmanEffectsVegetationSnow2008
@musselmanInfluenceCanopyStructure2012] with implications for the drying rates of litter fuels and
water stress throughout the fire season.

Previous studies in the western US suggest that many locations where aridity is increasing are also
experienced a reduction in spring snowpack that is correlated with both summer fuel moisture and
wildfire severity [@westerling.etal2006]. Despite the potential importance of changes in snow cover
to wildfire severity, the direct and indirect effects of snow cover and forest structure and the
interaction between these two factors on wildfire burn severity remain unclear. As climate warming
increases across the globe, and especially at northern latitudes, it is imperative to understand how
future changes in snow cover interact with stand-scale forest structure to influence wildfire burn
severity. Such information can greatly improve our knowledge as to the combined impacts of global
change and forest structure may contribute to wildfire severity. Additionally, this information can
inform forest management strategies for reducing the impacts of snow cover loss through structural
management regimes and helping understand wildfire hazard.

The primary goal of this study is to illustrate how forest structure and snow cover interact to
influence wildfire burn severity in the boreal forest of Ontario. We use structural equation
modelling to assess the strengths of the direct and indirect relationships between forest structure
and snow cover dynamics on wildfire burn severity. We fit models a different spatial scales to
compare how the strengths of these relationships differ across a climatic and forest structure
gradients. Finally, since snow cover dynamics are variable across a landscape

remote sensing

effect of scale and data assimilation

why do we study the scale we do, caveats?

# Methods

*Study Area*

Our study was focused on the boreal shield ecoregion of Ontario.

Between 2008 and 2018, over a third of the wildfires that burned in the boreal forest of Canada
occurred in the Ontario's boreal shield ecoregion [@coops.etal2018].

*Data*

**raw data**

Wildfires occurrence and polygon data were obtained for the years 2002 to 2019 from the Fire
Disturbance Area database created and openly distributed by the government of Ontario. We excluded
all fires that were classified as prescribed burns in the database. Fires that burn through a
previously burnt area, referred to as reburns, may burn more or less severely than fires that burnt
in areas without recent history of previous wildfire [@whitman.etal2019]. Therefore, we excluded all
reburns in our data following preexisting protocol outlined in [@whitman.etal2022]. Fires were
removed if the overlap between polygons were greater than the smallest area burned in our dataset
(40ha) and the gap between the fires was equal to or less than 26 years. We chose 26 years because
that represents the earliest year in which we had fire perimeter data for the first fire in our
dataset. After filtering out prescribed fires and reburns we had a total of 710 fires that burned
during 2002-2019.

**data acquisition, processing and assimilation**

Remotely sensed burn severity maps were obtained from Google Earth Engine (GEE) at a 30m spatial
resolution based on fire perimeters for wildfires that burned between 2002-2019. Burn severity was
measured using the relativized burn ratio (RBR). RBR is a Landsat-based fire severity metric that
represents the change in forest cover following a fire relative to the amount of pre-fire vegetative
cover rather than the absolute change. RBR is a more robust measure of burn severity comparing to
other Landsat-based approaches (e.g., rNBR) as it is more sensitive to pre-fire values
[@parks.etal2014]. Burn severity maps were generated from Landsat TM, ETM+ and OLI imagery following
a hybrid mean-compositing approach developed for the boreal region [@holsinger.etal2021] that
calculates the mean across a stack of pre-fire and post-fire imagery (see supplementary materials).

To determine snow free date (SFD) we used the global, 500-m spatial resolution, daily snow cover
product, version 6 from NASA's Moderate Resolution SpectroRadiometer (MODIS/Terra MOD10A1) to
generate maps in GEE for each fire perimeter using an algorithm developed for the northern
hemisphere [@crumley.etal2020]. Snow free date is the last date that snow is detected in a pixel,
calculated as the longest period without snow after a minimum of 5 days of snow cover. Maps were
produced for the entire extent of the Ontario boreal shield. We then calculated the mean SFD across
each fire event to produce an estimate of snow free date for each event. To produce an estimate of
snow free duration (SFDR) we calculated the mean julian day that each fire started to burn day
across the fire perimeter (julian day of burn). We then used the julian day of burn to calculate the
difference between mean day of burn and SFD. Mean julian day of burn was calculated as the julian
day of ignition averaged across the entire perimeter and was determined by modifying a previous
algorithm developed in GEE based on MODIS Terra and Aqua Thermal Anomalies and Fire Daily Global
datasets (MOD14A1 V6 and MYD14A1) to calculate the day that a fire stopped burning
[@holsinger.etal2021]. We acknowledge that there are potential biases in estimating snow cover from
satellite imagery [@hall2002, @parajka2008], however previous studies have used these data to
quantify snow free date in forest ecosystems [@gleason.etal2019].

We used remotely sensed forest structure diversity data that was produced at 250m spatial resolution
using MODIS imagery for the years 2001 and 2011 [@beaudoinMappingAttributesCanada2014]. In the
absence of stand-replacing disturbance, we assumed that canopy cover and biomass to be constant
between observation periods. Therefore, if the fire burned between 2002 and 2011 we assumed that
canopy cover and biomass values were equal to 2001 and for all fires after 2011 we assumed all
values were equal to 2011. For forest age, we used 2001 or 2011 as the reference age for all stands
and calculated the age of the stand at the given fire year based on the reference age.

We used terrain ruggedness index as a measure of topographic variation within each fire perimeter.
Terrain ruggedness index (TRI) is calculated using a digital elevation model and is the difference
in elevation (m) between one cell and the adjacent cell. Higher TRI values indicate more upland or
topographically complex areas whereas lower TRI values may indicate lowland areas or less
topographically complex areas. Topography is a strong driver of both wildfire severity
[@harris.taylor_2017] and snowmelt [@heldmyerInvestigatingRelationshipPeak2021]. Models evaluating
the effects of topography on severity and snowmelt have demonstrated a link between TRI and
variation in landscape patterns of both snowmelt and wildfire severity
[@heldmyerInvestigatingRelationshipPeak2021 @araya.etal2016]. We calculated mean TRI across each
fire perimeter from the Canadian Digital Elevation Model (CDEM) provided by Natural Resources
Canada.

To quantify the effects of weather on wildfire severity following snowmelt we used the drought code.
The drought code (DC) is a fire weather indices within the Canadian Forest Fire Behavior Prediction
system used to quantify drought and is a numeric rating of the average moisture content of deep and
compact organic layers [@chavardes.etal2019 @flanniganImpactsClimateChange2009 @girardin.etal2004a].
This code is a frequently used indicator of seasonal drought effects on forest fuels
[@girardin.wotton2009 @san-miguel.etal2020 @hanes.etal2020 @chavardes.etal2019 @girardin.etal2004a
@girardin.etal2004 @flannigan.etal2016] and the inclusion of this metric was twofold. First, it
allows us to control for potential drought effects and captures the effects of climate because it's
a cumulative index that is updated daily based on the previous days precipitation and temperature.
Secondly, the DC does not consider changes in moisture that occur do winter snow cover and begins at
an arbitrary value at the beginning of the snow free season. Because it is used by forest managers
across the globe to monitor wildfire risk, it may be a poor indicator of drought conditions,
especially in the early spring. Therefore, its inclusion allows us to compare the effects of SFD and
SFDR on wildfire severity in the presence of DC. We interpolated DC using thin plate splines to the
centroid of each fire perimeter from weather stations across our study region for the julian day of
burn.

In order to reduce (smooth) - signal fortin et al 2012

-   Because all remote sensing data aquired at different scales, we smooth data to the scale of the
    fire perimeter by calculating the median for all remote sensing data

**Data summary**

To address our objectives, we summarized burn severity by calculating the median and 90th percentile
RBR within each fire perimeter. We used median and 90th percentile RBR as the dependent variable in
our analyses. Median burn severity was used as measure of central tendency to describe the overall
trends in burn severity. We chose to measure the 90th percentile of burn severity within each
perimeter to represent burn severity extremes, which can be used to as a proxy to asses the drivers
of more high-severity fires. If the 90th percentile within a fire is high, it suggests that the fire
burned at a high-severity compared to a wildfire were the 90th percentile was lower. We referred to
the 90th percentile as burn severity extremes. The independent variables used our analyses were SFD,
SFDR, TRI, DC, stand age, canopy closure and total aboveground biomass. For stand structure, we
produced an estimate of stand age , canopy closure and total average biomass for each fire perimeter
by calculating the mean value of all pixels within an individual fire event. Likewise for SFD, we
calculated as the mean value of all pixels within a fire event and SFDR was calculated as the
difference between julian day of burn and snow free date. TRI was also average across the fire
perimeter.

We classified fires into three size classes based on previously defined thresholds of final fire
size for fires that burned in Canada's boreal forest [@stocks.etal2002]. Small fires were classified
as fires with a final size of \< 500 ha , medium fires where classified as fires with a final size
between 500 ha and 10,000 ha and large fires where all fires with a final size \> 10,000 ha.

Fires were classified into spring, summer and fall based on the julian day of burn. Fires where
julian day of burn was between April and June were classified as spring fires, following previous
studies in the boreal shield ecoregion of Ontario [@hanes.etal2020]. Summer fires were fires with a
julian day of burn in July or August, and fall fires were fires with a julian day of burn of
September or later.

**Statistical Analyses**

We examined the direct and indirect effects of snow free date, snow free duration and forest
structure on wildfire burn severity (median and 90th percentile burn severity) using piecewise
structural equation modeling (pSEM). Compared to traditional variance-covariance structural equation
modeling, piecewise approach is more flexible when dealing with problems that are common with
environmental data, such as multivariate normality and independence of errors [@lefcheck2016].

We began by specifying an a priori model of expected causal relationships between all variables
(Figure 1). We propose that forest structure represented by mean total aboveground biomass, mean
stand age and canopy closure within a fire perimeter will have direct effects on snow free date and
duration by increasing canopy interception, ablation, and sublimation. Yet, forest structure will
also influence wildfire burn severity. In addition, DC will influence wildfire burn severity and
snow free duration as drought increases the likelihood of observing an ignition in the boreal shield
of Ontario [@hanes.etal2020]. This model is subsequently a general test of mediation of forest
structure on the effects of SFD and SFDR on wildfire burn severity. We fit linear mixed effects
models using the nlme package in *R* [@pinheiroNlmeLinearNonlinear2023]. We fit our model with
random intercepts of both fire size class, and fire year (2001-2019) to control for fire size and
interannual effects. Extreme burn severity was left-skewed because it captures the 90th percentile
within each perimeter, to correct for this we square root transformed extreme burn severity to
reduce heteroscedascticity of the residuals. We evaluated multicollinearity by calculating the
variance inflation factor (VIF) on each set of regressions prior to fitting the SEM using a
threshold value of VIF \>0.8 (CITE). This approach resulted in no predictors being eliminated from
our model.

The strength of the direct, indirect effects and total effects were measured using the standardized
regression coeffiecents. Direct effects are quantified as the strength of the pathway between a
parent (i.e., cause) and its child (i.e., effect). Indirect effects are the product of all the
standardized regression coefficients for all indirect pathways traversed to get from a parent to a
child. Total causal effects are quantified as the sum of both direct and indirect pathways.

Overall model structure and goodness of fit was evaluated using Shipley's test of directed
separation [@shipley2009] and Fischer's C-statistic for pSEM [@shipley.douma2020]. Shipley's test of
directed separation tests whether the data can generate the hypothesized causal structure of the DAG
(figure 1), that is, whether the relationships specified in the DAG are supported by the data.
Directed separation tests whether a variable is conditionally independent from another variable
given a conditioning set. For example, whether... If test of directed separation in data are not
conditionally independent, it implies that it implies that the understanding of the causes in nature
as specified by our causal structure is wrong. Fischer's C-statistic is used to test whether you
hypothesized causal structure is due to sampling variation. If the C-statistic is greater than alpha
( $\alpha$ = 0.05 ) , you can accept your hypothesized structure as the observed differences could
reasonably occur due to sampling variation.

To assess how spatial and temporal scale influence the strength of the hypothesized causal
relationships we built models at different spatial and temporal scales and compared the strengths of
the standardized regression coefficients. We started with testing our hypothesized causal structure
on the entire boreal shield ecoregion of Ontario. Although there is significant east-west variation
in climate and weather patterns and fire regimes across the boreal shield of Ontario
[@beverly.martell2005 @hanes.etal2020 @hanes.etal2019], this allowed us to compare how relationships
change with scale, and whether certain relationships remain constant at different spatial scales.

After testing our structure across the entire boreal shield of Ontario, we split the our data into
east (n = ....) and west (n =...) boreal shield ecozones and examined how the strengths of the
relationships differed from east to west and across scales. To investigate potential causal
heterogeneity (i.e., whether the indirect effects of forest structure are different between or among
groups) in our hypothesized causal relationships given differences between ecozones we conducted a
multigroup analysis [@douma.shipley2021]. We initially fit a model for each ecoregion x response
combination where we allowed all path coefficients to be estimated as free parameters. We then fit
subsequent models for each indirect pathway from a forest structure variable (i.e., canopy closure,
biomass, stand age) to burn severity. In each model, we fixed the coefficients of all the paths
along the indirect pathways we were interested in by group (i.e., ecozone). For example, for the
indirect causal pathways between stand age and burn severity, we fixed all the paths from stand age
-\> snow free date -\> snow free duration and snow free duration -\> burn severity. To calculate
whether the paths differ by group (i.e., ecoregion) for each pathway of interest we summed both the
chi-square of each model and the degrees of freedom, and then calculated the difference between the
chi-square and degrees of freedom of the model where all paths are estimated freely and the fixed
model. We then tested the difference in chi-square and degrees of freedom against a chi-square
distribution. If the p-value is greater than $\alpha$ = 0.05 then we conclude that the indirect
pathway differs between groups.

To examine the effect of fire size on our hypothesize causal structure we split our data into three
groups based representing large, medium or small fires based on threshold assigned previously based
on [@stocks.etal2002]. Due to sample size limitations we conducted this analysis only for the subset
of fires that burned in the boreal shield west ecozone (n = ....). To compare differences in
hypothesized causal relationships we first fit models for both medium and extreme burn severity for
each fire size by initially leaving all paths free. We then preformed a multigroup analysis to
examine whether the indirect effects of forest structure are different between or among fire sizes.
**NOTE is the multigroup analysis examining diferences between seasons or ecoregions**

To examine the role of fire seasonality we split our fires into spring, summer and fall for the
entire boreal shield ecoregion and fit our initial models for each season and response (median or
extreme burn severity) combination. Since there is variation in fire seasonality across ecozones we
compared difference in the hypothesize causal relationships using a multigroup analysis where we
fixed the pathways by ecozone.

To test whether snow free date and snow free duration cause changes in the variation in burn
severity within an individual fire we measured the coefficient of variation of burn severity within
each perimeter. Therefore, this analysis examined whether the strengths of our hypothesized causal
relationships differed when the observed response was coefficient of variation of burn severity
within each fire perimeter. Since patches of high-severity fire are larger in large fires
[@cansler2014 @harvey.etal2016a], for this analysis we used only fires that were greater than 500
ha.

# Results

*Ontario boreal shield*

Across our entire study region, our structural equation model fit the data well with based on
good-of-fit static for both burn severity extremes (*C* = 14.177, *df* = 8 , *p* = 0.077) and median
burn severity (*C* = 14.177, *df* = 8 , *p* = 0.077) and revealed both direct and indirect
relationships between forest structural diversity, snow cover dynamics and burn severity. Our models
of median burn severity and burn severity extremes explained a high proportion of variance in both
snow free date *R*~2~ = 0.67 and snow free duration *R*~2~ = 0.84. Our models explained a higher
proportion of variance in burn severity extremes *R*~2~ = 0.20 relative to our models of median burn
severity *R*~2~ = 0.12.

We found a significant and negative direct relationship between total aboveground biomass and snow
free date (fig 1a., Table 1). Stand age had a significant but positive direct relationship on snow
free date (fig 1a, Table 1). We did not find a significant relationship between percent canopy
closure and snow free date (fig 1a, Table 1). Topography, represented by terrain ruggedness index,
had a significant and negative direct relationship on snow free date (fig 1a, Table 1). We found a
significant but negative direct relationship of snow free date on snow free duration (fig 1a, Table
1). We found significant but positive direct relationships of both drought and stand age on snow
free duration, with a stronger direct relationship between drought and snow free duration (fig 1a,
Table 1). Drought and snow free duration had significant direct relationships on median burn
severity and burn severity extremes (fig 1a, Table 1). Drought had a negative relationship with burn
severity extremes but a positive relationship with median burn severity (fig 1a, Table 1). The
relationship between snow free duration and median burn severity was negative whereas the
relationship between burn severity extremes and snow free duration was positive (fig 1a, Table 1).
All forest structure metrics had no significant direct relationship with median burn severity but
all metrics had a significant direct relationship with burn severity extremes (Table 1). Stand age
and canopy closure had a negative relationship with burn severity extremes whereas total aboveground
biomass have a positive relationship (fig1a).

The indirect relationship between forest structural diversity and median burn severity - calculated
by multiplying the standardized regression coefficients of the path coefficients - showed that the
effects of stand age were stronger than canopy closure but similar to that of total aboveground
biomass, although stand age was positive whereas biomass was negative (fig 1b). In contrast, the
direct effect of canopy closure on median burn severity is stronger than both stand age and total
aboveground biomass, both of which had relatively low direct effects (fig 1b). More importantly, the
total causal effects of total aboveground biomass and canopy closure were similar and much strong
than stand age (fig 1b).

The indirect effects of forest structural diversity on burn severity extremes were strongest for
stand age compare to both total aboveground biomass and canopy closure (fig 1a). The indirect
relationships for canopy closure and total aboveground biomass were positive whereas stand age had a
negative indirect relationship with burn severity extremes. Critically, the total causal effects of
total aboveground biomass were stronger compared to both stand age and canopy closure (fig 1a).
Total aboveground biomass and canopy closure had positive indirect relationships and stand age had a
negative indirect relationship.

The indirect relationship of drought were strongest for burn severity extremes compared to median
burn severity. Yet, drought had an indirect positive relationship with burn severity extremes and
indirect negative relationship for median burn severity (fig1a, fig1b). The total cause effects of
drought were positive. In contrast, topography had a positive indirect relationship with both median
burn severity and burn severity extremes, yet these relationships were weaker than drought. However,
because the indirect relationship of drought on burn median burn severity was strongly negative and
the direct relationship was strongly positive, the strength and direction of these effects cancel
each other out and the total causal effects of drought on median burn severity to be weaker than
topography (fig 1b).

*East vs. West Ontario shield*

For western Ontario, our SEM fit the data better compared to the models for all of Ontario based on
the goodness-of-fit statistic for both burn severity extremes (*C* = 9.864, *df* = 10, *p* = 0.453)
and median burn severity (*C* = 9.864, *df* = 10, *p* = 0.453) . Our model of median burn severity
and burn severity extremes explained the same proportion variance in snow free date *R*~2~ = 0.63
and snow free duration *R*~2~ = 0.83. Yet, our model of extreme burn severity explained a higher
proportion of variation in severity *R*~2~ = 0.20, compared to median burn severity *R*~2~ = 0.13.

For eastern Ontario, our SEM did not fit the data better compared to the model for all of Ontario
for extreme (*C* = 15.483, *df* = 10, *p* = 0.115) and median burn severity (*C* = 15.483, *df* =
10, *p* = 0.115). Similarly to western Ontario, the model of median burn severity and burn severity
extremes explained the same proportion variance in snow free date *R*~2~ = 0.78 and snow free
duration *R*~2~ = 0.87. Compared to western Ontario, our model explained more variance in both snow
free date and snow free duration. Yet, our model of extreme burn severity explained a higher
proportion of variation in severity *R*~2~ = 0.28, compared to median burn severity *R*~2~ = 0.18.

There was no significant direct relationship between biomass and snow free date in the western
boreal shield but we found a significant negative relationship in the eastern boreal shield (Table
\@ref(tab:S3) \@ref(tab:S4)\@ref(tab:S5) \@ref(tab:S6)) . We found a significant positive direct
relationship between stand age and snow free date in both ecoregions (Table \@ref(tab:S3)
\@ref(tab:S4)\@ref(tab:S5) \@ref(tab:S6)) . We found a significant negative direct relationship
between canopy closure and snow free date for the western boreal shield but a significant positive
relationship in the eastern boreal shield (Table \@ref(tab:S3) \@ref(tab:S4)\@ref(tab:S5)
\@ref(tab:S6)) . We found a significant negative relationship between topography and snow free date
in the west but not in the east (Table \@ref(tab:S3) \@ref(tab:S4)\@ref(tab:S5) \@ref(tab:S6)) . For
both ecoregions we found a significant negative relationship between snow free date and snow free
duration and significant positive relationship of drought and snow free duration (Table
\@ref(tab:S3) \@ref(tab:S4)\@ref(tab:S5) \@ref(tab:S6)) .

In the eastern boreal forest ecoregion, we found a significant negative direct relationship between
snow free duration and median burn severity but a significant positive relationship with burn
severity extremes (Table \@ref(tab:S4) \@ref(tab:S6)). We also found a significant negative direct
relationship between canopy closure and extreme burn severity (Table \@ref(tab:S4)).

Compared to the east, in the western boreal shield we found a significant positive direct
relationship between topography and median burn severity but not extreme burn severity (Table
\@ref(tab:S3) \@ref(tab:S5)). We found a significant positive direct relationship between snow free
duration and extreme burn severity but a significant negative relationship with median burn severity
(Table \@ref(tab:S3) \@ref(tab:S5)). We found significant positive relationship between drought and
both median and extreme burn severity (Table \@ref(tab:S3) \@ref(tab:S5)). We found significant
negative relationship of stand age and canopy closure and positive relationship of biomass on
extreme burn severity but no significant relationships with median burn severity(Table \@ref(tab:S3)
\@ref(tab:S5)).

Our multigroup analysis revealed that there is a difference from east to west for the indirect
pathway between biomass and extreme burn severity (*p* (x^2~L\\ML~^) = 2.2e-11) and the indirect
pathway between canopy closure and extreme burn severity (*p* (x^2~L\\ML~^) = 1.75e-11), but not
difference for the indirect pathway between stand age and extreme burn severity (*p* (x^2~L\\ML~^) =
0)(Figure \@ref(fig:fig3)). For median burn severity, we also observed a difference from east to
west for the indirect pathway from both biomass (*p* (x^2~L\\ML~^) = 2.89e-11) and canopy closure
(*p* (x^2~L\\ML~^) = 2.3e-11) but not for stand age (*p* (x^2~L\\ML~^) = 0)(Figure \@ref(fig:fig3)).

*Fire Heterogeneity*

For the entire boreal shield, test of directed separation revealed pathways between stand age and
snow free duration and biomass and snow free duration that were not in our proposed hypothesized
causal structure. We updated our model to include these paths. The updated causal model fit the data
better compared to both median and extreme burn severity at the scale of both the entire shield and
each ecoregion (*C* = 6.069, *df* = 6, *p* = 0.415). The updated model explained more variance in
snow free date *R*~2~ = 0.77 and snow free duration *R*~2~ = 0.90 than previous models. The updated
model explained more variation in heterogeneity in burn severity *R*~2~ = 0.29 compared to median
and extreme burn severity.

Stand age had a significant positive direct relationship with snow free date (*p* \<0.001), with an
increase in stand age causing snow free date to occur 29 days later (Table \@ref(tab:s7)). Biomass,
canopy closure and topography did not have significant direct relationships with snow free date
(Table \@ref(tab:s7)). Snow free date had a significant direct negative relationship with snow free
duration (*p* = 0.\<0.001), with an later snow free date decreasing duration by 34 days. Drought had
a significant positive direct relationship with snow free duration (*p* \<0.001), as drought
increased snow free duration increase by 67 days (Table \@ref(tab:s7)). Stand age had a significant
positive direct relationship with snow free duration (*p* = 0.\<0.001), with an increase in stand
age leading to longer snow free duration by 18 days (Table \@ref(tab:s7)). Biomass had a significant
negative direct relationship with snow free duration (*p* = 0.001), with an increase in biomass
causing snow free duration to be earlier by 11 days (Table \@ref(tab:s7)).

Snow free duration (*p* = 0.006) and snow free date (*p* = 0.028) had significant positive direct
relationships with heterogeneity in burn severity. Heterogeneity in burn severity increased by 32.8%
as snow free duration was longer, and increased by 19.5% as snow free date was later (Table
\@ref(tab:s7)). Topography had a significant negative direct relationship with heterogeneity in burn
severity (*p* \<0.001), with an increase in topography leading to a 26% decrease in heterogeneity
(Table \@ref(tab:s7)). Stand age had a significant positive direct relationship with heterogeneity
in burn severity (*p* =0.003), with an increase in stand age causing a 26% change in heterogeneity
(Table \@ref(tab:s7))

*Discussion*

Our results suggest that snow cover dynamics have a direct relationship with burn severity

**caveats**

-   scale of analysis

-   limitations of inference

-   remote sensing

Q1. How does stand structure and snow cover interact to influence wildfire burn severity in the
boreal shield?

Q2. do the strength of the direct and indirect relationships differ based on ecoregion and fire
seasonality?

Q3. Does snow cover influence within fire variability in burn severity?

# Figures

```{r, my dag}

```

```{r fig3, echo = FALSE, results = 'asis'}

```

# Tables

```{r s1, echo = FALSE, results = 'asis'}






```

```{r s2, echo = FALSE, results = 'asis'}
#

```

```{r S3, echo = FALSE, results = 'asis'}
#burn severity west/ext

```

```{r S4, echo = FALSE, results = 'asis'}
#burn severity east/ext}

```

```{r S5, echo = FALSE, results = 'asis'}
#burn severity west/med}
```

```{r S6, echo = FALSE, results = 'asis'}
#burn severity east/med}

```

```{r s7, echo = FALSE, results = 'asis'}
#heterogeneity full
```

```{r s8, echo = FALSE, results = 'asis'}
#heterogeneity multigroup
```
