---
title: "gather_clean_forest_structure_data"
author: "Jack Goldman"
date: "2023-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/jandr/Documents/PhD/Thesis/")

```


Load required packages 
```{r}
library(sf)
library(tidyverse)
library(ggplot2)
library(exactextractr)
library(raster)
```

# Extracting forest structure data for each fire event

This script gathers the NFI raster data that represent forest structure (age, biomass and crown closure)  2001 and 2011 at 250 m resolution for canada. To do this, I need to extract the avg for each fire using exact extract in a for loop and return as data frame.

To begin I load in all the rasters
```{r reporject rasters}
tabvg <- raster("C:/Users/jandr/Documents/PhD/Thesis/Data/NFI/Ontario_01/NFI_MODIS250m_2001_kNN_Structure_Biomass_TotalLiveAboveGround_v1_shield.tif")
tabvg_11 <- raster("C:/Users/jandr/Documents/PhD/Thesis/Data/NFI/Ontario_11/NFI_MODIS250m_2011_kNN_Structure_Biomass_TotalLiveAboveGround_v1_shield.tif")
stand_age_01 <- raster("C:/Users/jandr/Documents/PhD/Thesis/Data/NFI/Ontario_01/NFI_MODIS250m_2001_kNN_Structure_Stand_Age_v1_shield.tif")
stand_age_11 <- raster("C:/Users/jandr/Documents/PhD/Thesis/Data/NFI/Ontario_11/NFI_MODIS250m_2011_kNN_Structure_Stand_Age_v1_shield.tif")
crownClosure_01 <- raster("C:/Users/jandr/Documents/PhD/Thesis/Data/NFI/Ontario_01/NFI_MODIS250m_2001_kNN_Structure_Stand_CrownClosure_v1_shield.tif")
crownClosure_11 <- raster("C:/Users/jandr/Documents/PhD/Thesis/Data/NFI/Ontario_11/NFI_MODIS250m_2011_kNN_Structure_Stand_CrownClosure_v1_shield.tif")

#stack rasters
nfi.stack <- raster::stack(tabvg, tabvg_11, stand_age_01, stand_age_11, crownClosure_01, crownClosure_11)

names(nfi.stack) <- c("biomass_01", "biomass_11", "age_01", "age_11", "cc_01", "cc_11")

```

With the rasterstack we extract raster values to the geometry object.
```{r extract from NFI stack}
shpFires= st_read ("C:/Users/jandr/Documents/PhD/Thesis/OntBSdb/Fire_Perimeters/01_2020/OnShield_FirePerim_01to2020.shp")
#Calcs
shpFires = st_as_sf(shpFires)
shpFires = vect(shpFires)
shpFires = as(shpFires, "Spatial")

i = 1

datalist = list()
##extract nfi means
for (i in 1:nrow(shpFires)){
  

  shpFires1 <- shpFires[i,]#select a single fire perimeter
  extract1  <- exact_extract(nfi.stack, shpFires1, 'mean', full_colnames = TRUE, stack_apply = TRUE) # extract the mean
  extract1$Fire_ID <- shpFires1 %>% pull(Fire_ID) # get the specific fire id and make it a column
  datalist[[i]] <- extract1
}

df_nfi <- bind_rows(datalist) %>% 
  relocate(Fire_ID, .before = mean.biomass_01) %>% 
  rename(avgBio_01 = "mean.biomass_01", 
         avgBio_11 = "mean.biomass_11",
         age_01 = "mean.age_01",
         age_11 = "mean.age_11",
         cc_01 = "mean.cc_01",
         cc_11 = "mean.cc_11")

```


Lets divide the data based on the year assigned NFI year 2001 or 2011 - the rules are:
**For fires that burned between 2001-2011 assign nfi values from year 2001**
**For fires that burned between 2012-2020 assign nfi values from year 2011**

```{r}
df_nfi <- shpFires %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(Fire_ID,Fire_Year)) %>% 
  left_join(df_nfi, by = "Fire_ID") 


# For fires that burned between 2001-2011 assign nfi values from year 2001
df_nfi_01 <-  df_nfi %>% 
  filter(Fire_Year < 2012) %>% 
  dplyr::select(c(avgBio_01, age_01, cc_01, Fire_ID, Fire_Year))

# For fires that burned between 2012-2020 assign nfi values from year 2011
df_nfi_11 <-  df_nfi %>% 
  filter(Fire_Year >= 2012) %>% 
  dplyr::select(c(avgBio_11, age_11, cc_11, Fire_ID, Fire_Year))

```

Calculate stand age at time of fire. The rules are as follows:
Stand age at time of fire is assigned based on the survey that preceeds the fire year.
```{r}

#add a year based on the difference between 01 and last two digits of year
df_nfi_01 <- df_nfi_01 %>% 
  mutate(age = case_when(Fire_Year == "2001" ~ (age_01+0),
                         Fire_Year == "2002" ~ (age_01 +1),
                         Fire_Year == "2003" ~ (age_01 +2),
                         Fire_Year == "2004" ~ (age_01 +3),
                         Fire_Year == "2005" ~ (age_01 +4),
                         Fire_Year == "2006" ~ (age_01 +5),
                         Fire_Year == "2007" ~ (age_01 +6),
                         Fire_Year == "2008" ~ (age_01 +7),
                         Fire_Year == "2009" ~ (age_01 +8),
                         Fire_Year == "2010" ~ (age_01 +9),
                         Fire_Year == "2011" ~ (age_01 +10))) 

#add a year based on the difference between 11 and last two digits of year
df_nfi_11 <- df_nfi_11 %>% 
  mutate(age = case_when(Fire_Year == "2012" ~ (age_11+1),
                         Fire_Year == "2013" ~ (age_11 +2),
                         Fire_Year == "2014" ~ (age_11 +3),
                         Fire_Year == "2015" ~ (age_11 +4),
                         Fire_Year == "2016" ~ (age_11 +5),
                         Fire_Year == "2017" ~ (age_11 +6),
                         Fire_Year == "2018" ~ (age_11 +7),
                         Fire_Year == "2019" ~ (age_11 +8),
                         Fire_Year == "2020" ~ (age_11 +9))) 


```


clean up the final data:
1. remove age_01/11
2. remove extension(_..) for each column
```{r}
df_nfi_01 <-  df_nfi_01 %>% 
  dplyr::select(-c(age_01)) %>% 
  rename(avgBio = avgBio_01, 
         cc = cc_01)


#
df_nfi_11 <-  df_nfi_11 %>% 
  dplyr::select(-c(age_11))  %>% 
  rename(avgBio = avgBio_11, 
         cc = cc_11)
  
```


Combine the two dataframes
```{r}
df_nfi2 <-  df_nfi_01 %>% 
  bind_rows(df_nfi_11) %>% 
  relocate(Fire_ID, .before = avgBio) %>% 
  dplyr::select(-c(Fire_Year))

```

Save dataframe
```{r}
write.csv(df_nfi2, "SnowVariability_BurnSeverity/Data/forestStructure_01_2020.csv")
```

