---
title: "Dataset extraction and preparation"
author: "Jack A. Goldman"
date: "2023-01-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir ="C:/Users/jackg/Documents/PhD/Projects/SnowVariability_BurnSeverity/SnowVariability_WildfireSeverity/")
```

Load all the packages that are required for data gathering and tidying
```{r required packages}
library(tidyverse)
library(ggplot2)
library(sf)
library(exactextractr)
library(raster)

```

Our reference coordinates for our analysis will be the wildfire severity (rbr) maps for each fire.
To prep this we convert individual rbr rasters for each fire to a dataframe of the centroids of each pixel
and the corresponding value.

```{r calculate rbr and get x $ y coordinates}

path.In <- "C:/Users/jackg/Documents/PhD/Data/RBR_500m/rbr_500_clipped/"
rbr.list <- list.files(path.In, pattern = glob2rx("*rbr.tif"))


rbr.list = rbr.list[!grepl("_2001_", rbr.list)]



#create empty list

datalist = list()
##extract rbr means
for (i in 1:length(rbr.list)){
  
  #read in your raster
  ras.1 <- raster::raster(paste0(path.In,rbr.list[i]))
  fireID2 <- unlist(strsplit(rbr.list[i],"[.]"))[1] # get fire ID
  df <- raster::rasterToPoints(ras.1)
  df1 <- df %>%  as_tibble()
  df2 <- df1 %>%  mutate(fireID = fireID2)
  datalist[[i]] <- df2
  
  
}
#bind lists together and rename columns
df_rbr = datalist %>% bind_rows() %>%  
  rename(rbr = layer, lon = x, lat = y )
df_rbr$fireID <- df_rbr$fireID %>%   str_remove( "_rbr")

```

In order to extract predictors from other rasters at matching pixel locations we create square 
polygons around the centroid of each pixel using the reference resolution.
```{r create squares}
radius =  0.004491576


yPlus <- df_rbr$lat+radius
xPlus <- df_rbr$lon+radius
yMinus <- df_rbr$lat-radius
xMinus <- df_rbr$lon-radius

square=cbind(xMinus,yPlus,  # NW corner
	xPlus, yPlus,  # NE corner
	xPlus,yMinus,  # SE corner
	xMinus,yMinus, # SW corner
	xMinus,yPlus)

df_rbr = df_rbr %>% 
 add_column(num_ran = rep(1:133907),.after = "fireID") %>%
  unite("fireID2", c('fireID', 'num_ran'), sep = "_", remove = FALSE)
ID = df_rbr$fireID2



# create spatial polygons from coordinates
polys <- SpatialPolygons(mapply(function(poly, id) {
	  xy <- matrix(poly, ncol=2, byrow=TRUE)
	  Polygons(list(Polygon(xy)), ID=id)
	  }, 
	split(square, row(square)), ID),
	proj4string=CRS(as.character("+proj=utm +zone=11 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")))

plot(polys)
# Create SpatialPolygonDataFrame -- this step is required to output multiple polygons.
polys.df <- SpatialPolygonsDataFrame(polys, data.frame(id=ID, row.names = ID))

# as sf and remove rownames
polys.df2 = polys.df %>%  st_as_sf

rownames(polys.df2) <- NULL

#set crs
polys.df3 = polys.df2 %>% 
  st_transform(.,4326)


#join polygon to df-rbr
df_rbr = 
  df_rbr %>%  
  rename(id = "fireID2") %>% 
  left_join(polys.df3, by = "id") %>% 
  select(-c(id,num_ran))

#save polygons
st_write(df_rbr, "Data/severity_data_geoms.shp")

```


Create individual geometry object. This is a sf object that only contains the geometries.
```{r create geom object}
geom = df_rbr$geometry
geom = st_as_sf(geom)

```

Reproject NFI rasters to the desired resolution of the reference raster. We then create a rasterStack
of all NFI rasters.
```{r reporject rasters}
tabvg <- raster("C:/Users/jackg/Documents/PhD/Data/NFI_2001/Ontario/NFI_MODIS250m_2001_kNN_Structure_Biomass_TotalLiveAboveGround_v1_shield.tif")
tabvg_11 <- raster("C:/Users/jackg/Documents/PhD/Data/NFI_2011/Ontario/NFI_MODIS250m_2011_kNN_Structure_Biomass_TotalLiveAboveGround_v1_shield.tif")
stand_age_01 <- raster("C:/Users/jackg/Documents/PhD/Data/NFI_2001/Ontario/NFI_MODIS250m_2001_kNN_Structure_Stand_Age_v1_shield.tif")
stand_age_11 <- raster("C:/Users/jackg/Documents/PhD/Data/NFI_2011/Ontario/NFI_MODIS250m_2011_kNN_Structure_Stand_Age_v1_shield.tif")
crownClosure_01 <- raster("C:/Users/jackg/Documents/PhD/Data/NFI_2001/Ontario/NFI_MODIS250m_2001_kNN_Structure_Stand_CrownClosure_v1_shield.tif")
crownClosure_11 <- raster("C:/Users/jackg/Documents/PhD/Data/NFI_2011/Ontario/NFI_MODIS250m_2011_kNN_Structure_Stand_CrownClosure_v1_shield.tif")

#raster to reproject to
sdd_shield <- raster("C:/Users/jackg/Documents/PhD/Data/SDD2002Ontario_Shield.tif")

#stack rasters
nfi.stack <- raster::stack(tabvg, tabvg_11, stand_age_01, stand_age_11, crownClosure_01, crownClosure_11)

#reproject entire stack
nfi.stack500 = projectRaster(from = nfi.stack, to =sdd_shield,  method = "bilinear")

```

With the rasterstack we extract raster values to the geometry object.
```{r extract from NFI stack}
df_test = 
  exact_extract(nfi.stack500, geom , 'max',
                full_colnames = TRUE, stack_apply = TRUE)

#rename columns
df_test = df_test %>% 
  rename(tabvg_01 = "max.layer.1", 
         tabvg_11 = "max.layer.2", 
         age_01 = "max.layer.3",
         age_11 ="max.layer.4",
         closure_01 = "max.layer.5",
         closure_11 = "max.layer.6")

#join together
fireData = 
  df_rbr %>% 
  bind_cols(df_test)

#reorganise dataframe
fireData = 
  fireData %>% 
  relocate(geometry, .after = closure_11) %>% 
  relocate(lon, .after = closure_11) %>% 
  relocate(lat, .after = lon) %>% 
  relocate(fireID, .before = rbr)

#save it

```



```{r}
geom <-  st_read("data/severity_data_geoms.shp") %>% select(geometry, fireID, lat, lon) %>% rename(y = lat, x = lon) %>%   st_as_sf()
path_In <- "C:/Users/jackg/Documents/PhD/Data/NDMIperFire/NDMIperFire_clipped/"
ndmi_list <-  list.files(path_In, pattern = glob2rx("*_NDMI.tif"))

datalist = list()
for (i in 1:length(ndmi_list)) {
    r <- raster::raster(paste0(path_In, ndmi_list[i]))
    fireID <- unlist(strsplit(ndmi_list[i],"[.]"))[1] # get fire ID
    fireid <- gsub("*_NDMI", "", fireID)
    f_geom <- filter(geom, fireID == fireid)
    f_geom$values <- exact_extract(r, f_geom, "max", progress = F) 
    df <- st_drop_geometry(f_geom)
    datalist[[i]] <- df
}
df_ndmi = bind_rows(datalist)
df_ndmi = df_ndmi %>%  rename(c(ndmi = "values", lon = "x", lat = "y"))

#save it
write.csv(df_ndmi, "data/ndmi_per_pixel.csv")
```


```{r SDD}
# read in sdd files
sdd_list = list.files("C:/Users/jackg/Documents/PhD/Data/SDD/",pattern = "tif$", full.names = TRUE)
#create stack
sdd.stack = raster::stack(sdd_list)

#create list of names
sdd.names = paste0("sdd", sep = "_", rep(2:21))

#rename stack layers
names(sdd.stack) = sdd.names

# mask stack by shield ecoregion
shield = st_read("C:/Users/jackg/Documents/PhD/Data/Ontario_Boundaries/Ontario_Shield_v0.shp")
sdd.stack = mask(sdd.stack, shield )

#extract sdd values
df_sdd = 
  exact_extract(sdd.stack, geom , 'max',
                full_colnames = TRUE, stack_apply = TRUE)
# remove max
colnames(df_sdd) = str_remove(colnames(df_sdd), "max.")

# join to fireDate
fireData_v1 = 
  fireData %>% 
  bind_cols(df_sdd) %>% 
  relocate(geometry, .after = sdd_21) 


  
```

```{r scf}
# read in sdd files
scf_list = list.files("C:/Users/jackg/Documents/PhD/Data/SCF/",pattern = "tif$", full.names = TRUE)
#create stack
scf.stack = raster::stack(scf_list)

#create list of names
scf.names = paste0("scf", sep = "_", rep(2:21))

#rename stack layers
names(scf.stack) = scf.names

# mask stack by shield ecoregion
shield = st_read("C:/Users/jackg/Documents/PhD/Data/Ontario_Boundaries/Ontario_Shield_v0.shp")
scf.stack = mask(scf.stack, shield )

#extract sdd values
df_scf = 
  exact_extract(scf.stack, geom , 'max',
                full_colnames = TRUE, stack_apply = TRUE)
# remove max
colnames(df_scf) = str_remove(colnames(df_scf), "max.")

# join to fireDate
fireData_v2 = 
  fireData_v1 %>% 
  bind_cols(df_scf) %>% 
  relocate(geometry, .after = scf_21) 


#

```


```{r write data}
#save data with the geometries
st_write(fireData_v2, "fire_data_v2_geoms.shp")

#Remove geometries to save data without it
fireData_nogeom = fireData_v2 %>% 
  st_drop_geometry()

write.csv(fireData_nogeom, "data/fire_data_v2.csv")   

```

# Climate data gathering

```{r read in data}
clim_data = read.csv("C:/Users/jackg/Documents/PhD/Data/On_Fire_Elevation_1980-2020M.csv")
fireData = read.csv("data/fire_data_v2.csv") %>% select(-c(X))

```

*before we start, we want to create a year column in fire data*
```{r}
fireYear = read.csv2("C:/Users/jackg/Documents/PhD/Data/OnShield_FIRE_BS_v1.csv") %>% 
  select(c(raster_id, Fire_Year, FIRE_START)) %>% 
  rename(fireID = "raster_id") 

fireYear$fireID = str_remove(fireYear$fireID, "_rbr")

```


##  calculate sdd and scf for year of the fire

```{r}
#sdd/scf year of fire

# if year code first digit is one and nums first digit is one return matching col if not return other col
match <- function(nums, year_code){
  if (grepl("1", year_code)) {
  result <- nums[, grepl("1", colnames(nums))]
} else if(!grepl("1", year_code)) {
   result <- nums[, !grepl("1", colnames(nums))]
} 
  return(result)
} 

df_sdd_scf <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("fireID", "sdd_yof", "scf_yof")) #create an empty dataframe
for (i in 1:nrow(fireData)) {
fire_test = fireData[i,]
fire_test = fire_test %>%  
  separate(fireID,  c("ID", "Year", "Code"), "_") %>% 
  mutate(year_code = as.numeric(stringr::str_remove(Year, "20")))
year_code <- fire_test$year_code
nums <- fire_test[,names(fire_test) %in% colnames(fire_test)[grepl(year_code,colnames(fire_test))]]
nums_1 <- match(nums, year_code)
fire_best <- fire_test %>%  unite("fireID", ID:Code, sep = "_") %>% select(fireID)
colnames(nums_1) <- c("sdd_yof", "scf_yof")
input <- cbind(fire_best, nums_1)
new <- rep(input, ncol(df_sdd_scf)) 
df_sdd_scf[nrow(df_sdd_scf) + 1, ] <- new  
}

write.csv(df_sdd_scf, "Data/sdd_scf_yof.csv")

#issue with snow cover frequency

```

## what is the average sdd  and scf per fire

```{r}
#calculate average sdd per fire
average_sdd  = df_sdd_scf %>% 
  group_by(fireID) %>% 
  summarise(mean_sdd = mean(sdd_yof),
            mean_scf = mean(scf_yof)) 

write.csv(average_sdd, "Data/sdd_scf_perfire_yof.csv")



```



## lets clean up the climate data 
```{r tidy climate data}

# remove data prior to 2001, filter fire perims, get only wanted columns

clim_data$ID1 <- str_remove(clim_data$ID1, "_rbr") # remove rbr at the end of id

average_sdd <- average_sdd %>%  rename_all(tolower) # change all colnames in average sdd to lowercase



clim_data_v1 <- 
  clim_data %>% 
  filter(Year >= 2001) %>% # filter out climate data from years before 2001
  rename(fireID = "ID1") %>% # rename ID1 column to fireID
  semi_join(fireData, by = "fireID") %>% # filter out fires that did no burn btw 01-2020
  left_join(fireYear, by = "fireID") %>% # join fire related metadata
  relocate(Fire_Year, .after = "fireID") %>% # relocate for visualization ~~~ not necessary
  relocate(FIRE_START, .after = "Fire_Year") %>% # relocate for visualization ~~ not necessary
  select(-c(ID2, Longitude, Latitude, Elevation)) %>% # remove unnecessary columns
  mutate(fire_month = month(ymd(FIRE_START)), .after = Fire_Year, #get the mont that the fire burned # not necessary
         fire_julday = yday(ymd(FIRE_START))) %>% # get julday for calendar year calculation
  rename_all(tolower) %>% 
  select(c(year:fire_start,tave01:tave12, cmi01:cmi12, rh01:rh12, ppt01:ppt12)) %>% # select important variables
  mutate(matches = case_when(
    year == fire_year ~ "yes",
    TRUE ~ "no")) %>% # create matches for the years we need
  filter(matches == "yes") %>% 
  select(-c(matches, year)) %>% 
  left_join(average_sdd, by = "fireid") %>% 
  relocate(c("mean_sdd", "mean_scf"), .after = fire_start) %>% 
  mutate(caly_sdd = (mean_sdd - 81), .after = mean_scf,
         sdd_date = as.Date(caly_sdd, origin = as.Date("1970-01-01"))) %>% 
  separate(sdd_date,  c("sdd_y", "sdd_m", "sdd_d"), "-") %>%
    separate(fire_start,  c("f_y", "f_m", "f_d"), "-") %>% 
  select(-c(f_y, f_d, sdd_y, sdd_d, mean_sdd, mean_scf, caly_sdd, fire_month, fire_julday)) 
  


```


## lets calculate climate from time since snowmelt to fire month

```{r calculate average climate variables between sdd month and fire month}
#create empty dataframe
df_fullclim <-  setNames(data.frame(matrix(ncol = 4, nrow = 0)), c("fireID", "avg_cmi", "avg_ppt", "avg_vpd"))

#load the vpd calc function
source("scripts/vpd_function.R")

#loop that calculates avg cmi, ppt and vpd
for (i in 1:nrow(clim_data_v1)) {
  #read in data
  clim <-  clim_data_v1[i,] 
  
  #cmi calc
  cmi_v1 <-  clim %>%  select(c(starts_with("cmi"), sdd_m, f_m, fireid))
  sdd_code <- cmi_v1$sdd_m
  f_code <- cmi_v1$f_m
  to_match <- c(sdd_code:f_code)
  nums <- cmi_v1[,names(cmi_v1) %in% colnames(cmi_v1)[grepl(paste(to_match,collapse="|"),colnames(cmi_v1))]]
  results <- lapply(nums,as.numeric)
  nums <- as.data.frame(do.call(cbind, results))
  avg_nums <- nums %>% mutate(avg = rowMeans(nums)) %>%  select(avg)
  avg_nums <- round(avg_nums, digits = 2)
  
  #ppt calc
  ppt_v1 <-  clim %>%  select(c(starts_with("ppt"), sdd_m, f_m, fireid))
  sdd_code_p <- ppt_v1$sdd_m
  f_code_p <- ppt_v1$f_m
  to_match_p <- c(sdd_code_p:f_code_p)
  nums_p <- ppt_v1[,names(ppt_v1) %in% colnames(ppt_v1)[grepl(paste(to_match_p,collapse="|"),colnames(ppt_v1))]]
  results_p <- lapply(nums_p,as.numeric)
  nums_p <- as.data.frame(do.call(cbind, results_p))
  avg_nums_p <- nums_p %>% mutate(avg = rowMeans(nums_p)) %>%  select(avg)
  avg_nums_p <- round(avg_nums_p, digits = 2)
  
  #vpd calc
  tave_v1 <-  clim %>%  select(c(starts_with("tave"), sdd_m, f_m, fireid))
  rh_v1 <-  clim %>%  select(c(starts_with("rh"), sdd_m, f_m, fireid))

  sdd_code_t <- tave_v1$sdd_m
  f_code_t <- tave_v1$f_m
  to_match_t <- c(sdd_code_t:f_code_t)
  sdd_code_r <- rh_v1$sdd_m
  f_code_r <- rh_v1$f_m
  to_match_r <- c(sdd_code_r:f_code_r)
  
  nums_t <- tave_v1[,names(tave_v1) %in% colnames(tave_v1)[grepl(paste(to_match_t,collapse="|"),colnames(tave_v1))]]
  nums_r <- rh_v1[,names(rh_v1) %in% colnames(rh_v1)[grepl(paste(to_match_r,collapse="|"),colnames(rh_v1))]]
 
  results_t <- lapply(nums_t,as.numeric)
  results_r <- lapply(nums_r,as.numeric)

  
  nums_vpd_1 <- as.data.frame(do.call(rbind, results_t)) # i need to cbind so two columns one with tave and one with rh
  nums_vpd_2 <- as.data.frame(do.call(rbind, results_r)) # i need to cbind so two columns one with tave and one with rh
  nums_vpd <-  cbind(nums_vpd_2, nums_vpd_1)
  colnames(nums_vpd) <- c("rh", "tave")
  
  avg_vpd_pm <- nums_vpd %>% 
    mutate(vpd_m = vpd(rh, tave)) %>% 
    select(-c(rh, tave)) %>% 
    rownames_to_column("month") %>% 
    pivot_wider(names_from = month, 
                values_from = vpd_m) 
  
  avg_vpd <- avg_vpd_pm %>% mutate(avg_vpd = rowMeans(avg_vpd_pm)) %>% select(avg_vpd)
  
  #dataframe combine
  clim_v2 <- clim_v1 %>% select(fireid) %>% rename(fireID = "fireid")
  colnames(avg_nums) <- c("avg_cmi")
  colnames(avg_nums_p) <- c("avg_ppt")
  input <- cbind(clim_v2, avg_nums, avg_nums_p, avg_vpd)
  new <- rep(input, ncol(df_fullclim)) 
  df_fullclim[nrow(df_fullclim) + 1, ] <- new  
}

#save result
write.csv(df_fullclim, "data/timesincesnowmelt_climate_averages.csv")
```



## combine all datasets

```{r}
fireData <-  fireData %>% 
  select(-c(12:51))
fireData <- fireData %>% 
  bind_cols(df_sdd_scf)

firedata_v3 <- fireData %>% 
  bind_cols(df_fullclim)

```   

