---
title: "full_clean_data"
author: "Jack Goldman"
date: "2023-02-28"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/jandr/Documents/PhD/Thesis/")

```

Load required packages

```{r}
library(tidyverse)
```

Load in data

```{r}
df_snow <-  read.csv("Projects/snow_cover_dynamics/SWE_Overwinter_DroughxBurn_Severity/Data/OnShield_SnowPhenology_01to2020_v1.csv")
df_burnday <- read.csv("SnowVariability_BurnSeverity/Data/BurnDay_01_2020.csv")
df_doPS <- read.csv("SnowVariability_BurnSeverity/Data/dayPeakSnow_sweMax_01_2020.csv")
df_climavgs <-  read.csv("Data/01_2020/timesincesnowmelt_climate_averages.csv")
df_fs <- read.csv("SnowVariability_BurnSeverity/Data/forestStructure_01_2020.csv")
```

clean data frames

```{r}
#change raster_id to fire id and remove _rbr
df_snow <-  df_snow %>% 
  rename(Fire_ID = raster_id) %>% 
  mutate(Fire_ID = str_remove(Fire_ID, "_rbr")) %>% 
  select(-c(X))

#clean day of peak snow, remove x column
df_doPS <- df_doPS %>% 
  select(-c(SweMax, X))

#clean burnday, remove x column
df_burnday <- df_burnday %>% 
  select(-c(X))

```

Join burn day and doPS to snow phenology dataframe

```{r}
df_snow <- df_snow %>% 
  left_join(df_burnday, by = "Fire_ID") %>% 
  left_join(df_doPS, by = "Fire_ID")
```

Calculate time from peak snow to melt day - time to disapearance following peak snow (tdps) & Calculate time since Snow Melt (tssm)

```{r}
df_snow <- df_snow %>% 
  mutate(tdps = (SnowMeltDay - doPS), 
         tssm = (BurnDay - SnowMeltDay))
```

Attach montly clim

```{r}
df_snow <-  df_snow %>% 
  left_join(df_fullclim, by = "Fire_ID")
```

For GEE calculations lets turn the julian dates to y/m/d. Makes it easier when working with dates in other programs. To do so we create a function and which we will then map trough the columns we need

```{r}
#this function using the date origin and the jul date to compute y/m/d
comb_ymd <- function(origin, StartBurnDay){
  datee = as.Date(as.numeric(StartBurnDay), origin = origin) 
  return(datee) 
}
```

create fire start day in y/m/d

```{r}


df_snow <-  df_snow %>% 
  mutate(origin = as.Date(ISOdate(Fire_Year, 1, 1))) %>% 
  mutate(Fire_Start = map2(origin, BurnDay , comb_ymd)) %>% 
  unnest(Fire_Start) %>% 
  select(-c(origin)) %>% 
  relocate(Fire_Start, .after = Fire_Year) %>% 
  relocate(BurnDay, .before = SnowAccDay)
```

create snow Melt Day in y/m/d

```{r}
df_snow <- df_snow %>% 
  mutate(origin = as.Date(ISOdate(Fire_Year, 1, 1)),
         Snow_Melt = map2(origin, SnowMeltDay , comb_ymd)) %>% 
  unnest(Snow_Melt) %>% 
  select(-c(origin)) %>% 
  relocate(Snow_Melt, .before = SnowCoverDays)
```

create Snow Acc Day in y/m/d

```{r}
df_snow <- df_snow %>% 
  mutate(Fire_Year_pre = (Fire_Year - 1),
         origin = as.Date(ISOdate(Fire_Year_pre, 1, 1)),
         Snow_Acc = map2(origin, SnowAccDay , comb_ymd)) %>% 
  unnest(Snow_Acc) %>% 
  select(-c(origin, Fire_Year_pre)) %>% 
  relocate(Snow_Acc, .before = SnowAccDay)
```

create peak snow day in y/m/d

```{r}
df_snow <- df_snow %>% 
  mutate(origin = as.Date(ISOdate(Fire_Year, 1, 1)),
         ymdPS = map2(origin, doPS, comb_ymd)) %>% 
  unnest(ymdPS) %>% 
  select(-c(origin)) %>% 
  relocate(ymdPS, .before = peakSnow)
```

join forest structure

```{r}
df_snow <- df_snow %>% 
  left_join(df_fs , by = "Fire_ID") %>% 
  dplyr::select(-c(X))
```

save it

```{r}
write.csv(df_snow, "SnowVariability_BurnSeverity/Data/snow_fire_dynamics.csv")
```

join topography - after joining topography we save a new separate file because of we dont want Tri that equalled NA

```{r}


df_snow <- df_snow %>% 
  left_join(df_tri2, by = "Fire_ID") 


```

Add RBR to the dataframe: split into fire size classes based on - [@stocks2002] small (200--500 ha), medium (500--10 000 ha), and large (\>10 000 ha)

```{r}

df_rbr <- read.csv("C:/Users/jandr/Documents/PhD/Thesis/OntBSdb/Clean_Tables/OnShield_Fire_BS_v2.csv")

df_rbr <- df_rbr %>% 
  dplyr::select(raster_id, RBR_median, RBR_quant, RBR_coefvar, FIRE_FINAL) 

df_rbr$raster_id <- str_remove(df_rbr$raster_id, "_rbr")

df_rbr <- df_rbr %>% 
  rename(Fire_ID = raster_id)

df_snow <- df_snow %>% 
  left_join(df_rbr, by = "Fire_ID")

#create dummy variable

df_snow <- df_snow %>% 
  mutate(size_class = case_when(
    FIRE_FINAL <= 500 ~ 1, 
    FIRE_FINAL <= 10001 & FIRE_FINAL >= 501 ~ 2,
    TRUE ~ 3
  ))

write.csv(df_snow, "SnowVariability_BurnSeverity/Data/snow_fire_dynamics_final.csv")


```

Add ppt and tmax to the dataframe

```{r}
df_ppttmax <- read.csv("SnowVariability_BurnSeverity/Data/ppt_tmax_timesincemelt_01_2020.csv")

df_snow <- df_snow %>% 
  left_join(df_ppttmax, by = "Fire_ID")

df_snow <- df_snow %>% 
  dplyr::select(-c(X.y, avg_vpd, avg_cmi, X.x))

write.csv(df_snow, "SnowVariability_BurnSeverity/Data/snow_fire_dynamics_final.csv")

```
