---
title: "moisture_indices_interpolation"
author: "Jack Goldman"
date: "2023-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/jandr/Documents/PhD/Thesis/")

```





**FINAL DATA CONSISTS OF:**
      **-Years 2001 to 2019**
      **-Boreal Shield of Ontario**
      **-Months May to September**


Load required packages
```{r}
library(lubridate)
library(sf)
library(gstat)
library(dplyr)
library(fields)
library(sp)

```


Load required data - we need both the Fire Perimeter data with Start Date and the FWI Data for 1980 to 2019
```{r}
df_wx <- read.csv("Data/DailyFWI_80to2019/daily_wx_1980_2019.csv")
shp_fire <- st_read("OntBSdb/Fire_Perimeters/01_2020/OnShield_FirePerim_01to2020.shp") # has all reuiqred infor
df_fire <- read.csv("OntBSdb/Topography/On_Fire_Elevation.csv") # as all centroids
df_burnday <- read.csv("Data/01_2020/boreal_fires_start.csv")

```


Join all the required info from the shp_fires to data with the fire centroids (elevation data)
```{r}
shp_fire %>% glimpse()

#select required columns
shp_fire <- shp_fire %>% 
  select(Fire_ID, FIRE_START, Fire_Year)

# rename id columns to match both dfs and join shpfire to df_fire
df_fire <- df_fire %>% 
  rename(Fire_ID = ID1) %>% 
  mutate(Fire_ID = str_remove(Fire_ID, "_rbr")) %>% 
  left_join(shp_fire, by = "Fire_ID") 



```

Get fire start day and join to df_fire

```{r}
df_burnday %>% 
  glimpse()

# convert julday in StartBurnDay column to ymd, need to specify year 

create_y <- function(origin, StartBurnDay){
  datee = as.Date(as.numeric(StartBurnDay), origin = origin) 
  return(datee) 
  }
  
df_burnday2 <- df_burnday %>% 
  mutate(filter_burned = yday(ymd(StartBurnDay)),
         origin = as.Date(ISOdate(Fire_Year, 1, 1))) %>% 
  filter(is.na(filter_burned)) %>%  
  mutate(burned = map2(origin, StartBurnDay , create_y)) %>% 
  unnest(burned) %>% 
  select(-c(StartBurnDay, origin, filter_burned)) %>% 
  rename(StartBurnDay = burned) 

# attach df_burnday2 back to df_burnday
df_burnday <- df_burnday %>% 
  mutate(filter_burned = yday(ymd(StartBurnDay)),
         origin = as.Date(ISOdate(Fire_Year, 1, 1)),
         StartBurnDay = ymd(StartBurnDay)) %>% 
  filter(!is.na(filter_burned)) %>% 
  select(-c(filter_burned, origin)) 

  
df_burnday <-  df_burnday %>% bind_rows(df_burnday2)

#join fire start day to df_fire
df_burnday <- df_burnday %>% select(Fire_ID, StartBurnDay)
df_fire <- df_fire %>% left_join(df_burnday, by = "Fire_ID")  


# remove 
df_fire  <- df_fire %>% select(-c(ID2, geometry))
#remove uneeded dfs
rm(df_burnday2, df_burnday, shp_fire)
```


I want to get 
```{r}
##Format weather date
df_wx$WX_DATE <- ymd(df_wx$WX_DATE)
#create interpolation date
df_fire$interp_date <- as.Date(df_fire$StartBurnDay)

```

remove fires that burned in 2020 and that burned in may or september
```{r}
df_fire <-  df_fire %>% 
  filter(Fire_Year <= "2019") 

## Only keep fires from May to September 
df_fire$month <- month(df_fire$interp_date)
## 756 to 748
df_fire <- subset(df_fire, df_fire$month > 4 & df_fire$month < 10)
```



TPS interpolation
```{r}
## Convert fire data to a SPDF based on lat/long locations 
coordinates(df_fire) <- ~ long + lat

##Get unique fire date 
x <- unique(df_fire$interp_date)

##Create empty dataframe to dump results in 
final_df <- NULL


for (j in (x)){
  #subset fire id
  k <- as.Date(j, origin = "1970-01-01")
  
  f <- df_fire[df_fire$interp_date == k,]
  
  w <- df_wx[df_wx$WX_DATE == k,]
  
  xx <- coordinates(f)
  
  inner_df <- NULL
  

  ##For every day, interpolate ISI, FWI, BUI
  for (i in list("DC", "FFMC", "DMC")){
    
    df_sub <- w[,c("LONGITUDE", "LATITUDE", "WX_DATE",i)]
    
    
    ##Create spline that takes long/lat and weather variable 
    spln <- Tps(df_sub[,1:2], df_sub[,i])
    
    ## Predict ISI, FWI, and BUI based on fire coordinates 
    pred <- predict(spln, x=as.matrix(xx) )
    
    
    df2 <- data.frame(cbind(xx, pred))
    
    
    if (i == "DC"){
      inner_df <-df2
      colnames(inner_df[3]) <- c("DC_pred")
    } else if (i == "DMC"){
      inner_df$DMC_pred <- df2$V3
    } else if (i == "FFMC"){
      inner_df$FFMC_pred <- df2$V3
    } 
    inner_df$DATE <- paste0(k)
  }
  final_df <- rbind(final_df, inner_df)
}

write.csv(final_df, "SnowVariability_BurnSeverity/Data/wx_interp_daily.csv")



## Rename ISI (did not work)
names(final_df)[names(final_df) == "v3"] <- "DC_pred"

## Change date in final_df to date format
final_df$DATE <- ymd(final_df$DATE)

```

Interpolation Cleaning - because FFMC and DMC values are bounded (e.g., 1-93 for FFMC), TPS splines do not know this and 
as a consequence the data could has value above or below these values, we need to control for this setting all values that fall outside of the possible range as the avg value for that day across all sites.
```{r}
##Completed observation dataset
ffmc <- df_wx %>% group_by(WX_DATE) %>% summarise(min.ffmc=min(FFMC), max.ffmc=max(FFMC))
##Join min/max BUI of completed data with interpolated dataset
final_df <- left_join(final_df, ffmc, by = c("DATE" = "WX_DATE"))
##Change the values so they are not outside the min and max. 
final_df$ffmc2 = case_when(final_df$FFMC_pred < final_df$min.ffmc ~ final_df$min.ffmc,
                           final_df$FFMC_pred > final_df$max.ffmc ~ final_df$max.ffmc,
                           TRUE ~ final_df$FFMC_pred)

############ DMC #############

##Completed observation dataset
dmc <- df_wx %>% group_by(WX_DATE) %>% summarise(min.dmc=min(DMC), max.dmc=max(DMC))
##Join min/max BUI of completed data with interpolated dataset
final_df <- left_join(final_df, dmc, by = c("DATE" = "WX_DATE"))
##Change the values so they are not outside the min and max. 
final_df$dmc2 = case_when(final_df$DMC_pred < final_df$min.dmc ~ final_df$min.dmc,
                           final_df$DMC_pred > final_df$max.dmc ~ final_df$max.dmc,
                           TRUE ~ final_df$DMC_pred)
```



Clean up the final dataset, remove predicted values for FFMC and DMC and keep corrected values.
```{r}

final_df2 <- final_df %>% 
  select(c("dmc2", "ffmc2", "DC_pred", "DATE", "lat", "long")) %>% 
  rename(dmc = "dmc2", ffmc = "ffmc2", dc = "DC_pred") %>% 
  rename_all(tolower)


```

Merge fire and weather data 
```{r}
## Convert fire data back to data frame
df<- df_fire %>% 
  as.data.frame() %>% 
  left_join(final_df2, by = c("interp_date" = "date", "long" = "long", 
                                           "lat" = "lat"))

write.csv(df, "SnowVariability_BurnSeverity/Data/FWI_moisture_codes_01_19.csv")

```

