---
title: "sem"
author: "Jack Goldman"
date: "2023-03-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/jgoldman/Work/PhD/forest-snow-fire-interactions/")

```

## Data analysis for structural equation Model

Load packages

```{r}
library(piecewiseSEM)
library(nlme)
library(tidyverse)

```

Read in data

```{r}
df_snow <- read.csv("Data/0-forest-snow-fire-data.csv") %>% 
  filter(Fire_Year < 2020 | Fire_Year <=2001) 

#where are nas
nas <-  df_snow %>% 
  select_if(function(x) any(is.na(x))) #ecoregion na's

#omit na's
new_data <- df_snow %>% select(-c(ecoregion)) %>% na.omit(.)

#make sure DC is greater than 0
new_data <- new_data %>% filter( dc > 0) 



```

RBR quant is negatively skewed. Fix this using square root
transformation for negatively skewed data

```{r}
#sqrt transform rbr quant
new_data <- new_data %>% mutate(rbr_qs = sqrt((max(RBR_quant+1)-RBR_quant)))
hist(new_data$rbr_qs)
```

*Notes:* Fit structural equation model - all fires: - I removed climate
data because DC takes into consideration ppt and temperature from the
first day of the fire season. - Should snow Accumulation day be included
because of Chelene's past work? Chelene found that Overwintering DC has
an effect, therefore the day that snow fell would determine
overwintering DC. -DC is correlated with time since snow melt because DC
is cumulative and would differ based on the amount of days between the
start of the fire season and fire burn, should theoretically increase as
TSSM increases - TSSM includes burn day, therefore ignition
probabilities are likely influenced by canopy closure, bioamass and age.
As biomass increases, ignition probability is higher, and as canopy
closure increases, ignition and spread is more ikely and as age
increases, more flammable fuel.

Get month/season of fire of fire

```{r}


# fire seasonality

#create season
new_data <- new_data %>% 
  mutate(fmonth = lubridate::month(Fire_Start),
         season = case_when(fmonth <= 6 ~ 1,
                            fmonth > 6 & fmonth <=8 ~ 2,
                          fmonth > 8 ~ 3))

#east vs. west
new_data_w <- new_data %>% filter(pyroregion == 1)
new_data_e <- new_data %>% filter(pyroregion == 0)

#spring fires
new_data_sp <- new_data %>% filter(season == 1) # 195 observations
west_sp <- new_data_w %>% filter(season == 1) # 126 observations
east_sp <- new_data_e %>% filter(season == 1) # 69 observations 

#summer fires
new_data_sm <- new_data %>% filter(season == 2) # 412 observations
west_sm <- new_data_w %>% filter(season == 2)# 297 observations
east_sm <- new_data_e  %>%  filter(season ==2)# 115 observations

#fall fires
new_data_fl <- new_data %>% filter(season == 3) # 82 observations
west_fl <- new_data_w %>% filter(season == 3)# 71 observations
east_fl <- new_data_e  %>%  filter(season ==3)# 11 observations


```

Create a VIF function that will allow me to test multicollinearity on
linear mixed effects (nlme) models

```{r, vif lme function}
vif.lme <- function (mod) #taken from github/gmonette/spida15
{
    if (any(is.na(fixef(mod)))) 
        stop("there are aliased coefficients in the model")
    v <- vcov(mod)
    mm <- model.matrix(formula(mod), mod$data)
    assign <- attributes(mm)$assign
    if (names(fixef(mod)[1]) == "(Intercept)") {
        v <- v[-1, -1]
        assign <- assign[-1]
    }
    else warning("No intercept: vifs may not be sensible.")
    terms <- labels(terms(mod))
    n.terms <- length(terms)
    if (n.terms < 2) 
        stop("model contains fewer than 2 terms")
    R <- cov2cor(v)
    detR <- det(R)
    result <- matrix(0, n.terms, 3)
    rownames(result) <- terms
    colnames(result) <- c("GVIF", "Df", "GVIF^(1/2Df)")
    for (term in 1:n.terms) {
        subs <- which(assign == term)
        result[term, 1] <- det(as.matrix(R[subs, subs])) * det(as.matrix(R[-subs, 
            -subs]))/detR
        result[term, 2] <- length(subs)
    }
    if (all(result[, 2] == 1)) 
        result <- result[, 1]
    else result[, 3] <- result[, 1]^(1/(2 * result[, 2]))
    result
  }
```

Test individual model pathways for multicollinearity using VIF function

```{r, Extreme burn severity}

#cc and avg biomass are highly correlated
vif.lme(lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
    avgBio + cc, random = list(size_class =~1, Fire_Year =~1), data = new_data))

vif.lme(lme(tssm ~ sdd + tri , random = list(size_class =~1, Fire_Year =~1), data = new_data))

#cc and avg biomass are highly correlated
vif.lme(lme(sdd ~ cc + avgBio + age + tri , random = list(size_class =~1, Fire_Year =~1), data = new_data))

#compare models with cc or avgbio AIC 
rbr_bio <- lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
    avgBio,random = list(size_class =~1, Fire_Year =~1), data = new_data)

rbr_cc <- lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
    cc   , random = list(size_class =~1, Fire_Year =~1), data = new_data)
AIC(rbr_bio, rbr_cc) # CC is greater by .400 , I chose to use bio because it better predicts sdd

sdd_bio <- lme(sdd ~  avgBio + age + tri , random = list(size_class =~1, Fire_Year =~1), data = new_data)

sdd_cc <- lme(sdd ~  cc+ age + tri , random = list(size_class =~1, Fire_Year =~1), data = new_data)

AIC(sdd_bio, sdd_cc)# biomass has lower AIC




```

```{r, Median Burn Severity}
#cc and avg biomass are highly correlated
vif.lme(lme(RBR_median ~ dc + tssm + sdd +tri +age + 
    avgBio + cc, random = list(size_class =~1, Fire_Year =~1), data = new_data))

#compare models with cc or avgbio AIC 
mean_bio <- lme(RBR_median ~ dc + tssm + sdd +tri +age + 
    avgBio,random = list(size_class =~1, Fire_Year =~1), data = new_data)

mean_cc <- lme(RBR_median ~ dc + tssm + sdd +tri +age + 
    cc   , random = list(size_class =~1, Fire_Year =~1), data = new_data)
AIC(mean_bio, mean_cc) # CC is greater by 1.300 , I chose to use bio because it better predicts sdd

```

# All of Ontario

### Extreme

```{r, Extreme all of Ontario}
sem_ext <- psem(
  lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
         avgBio, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
  lme(tssm ~ sdd + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
  lme(sdd ~  avgBio + age + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML")
)


#summary
summary(sem_ext)
```

There was an significant independence claim which was that drought code
has a significant impact on time since snowmelt. We refit the model to
include the pathway tssm \<- dc.

```{r, Extreme all of Ontario refit final model}
sem_ext_2 <- psem(
  lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
        avgBio , random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
  lme(tssm ~ sdd + tri + dc, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
  lme(sdd ~  avgBio + age + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML")
)

#summary
summary(sem_ext_2)

lapply(sem_ext_2, function(i) hist(resid(i))) # looks good

```

SDD did not have an impact on extreme burn severity - this suggests that
snowmelt timing does not influence burn severity, a possible explanation
for this is that snowmelt timing is heterogeneous across a landscape and
by averaging snowmelt timing and rbr across the perimeter we are
homogeneize both SDD and RBR, therefore we are unable to pick up
variations in both sdd/rbr. Further investigations that consider
variation in both. TSSM had a significant impact on extreme burn
severity - As time sine snowmelt increased, burn severity increased. DC
had a significant impact on extreme burn severity - As DC increased,
burn severity decreased. Forest structural diversity was a strong
predictor of extreme burn severity: - as age increased, burn severity
decreased - older stands decreased burn severity. - as biomass
increased, burn severity increased - overall biomass increased burn
severity. - as canopy closure increased burn severity decreased -
suggesting that live foliage biomass significantly reduces burn
severity, this may be a result of reduced flammability of live foliage.
SDD had a significant impact on time since snowmelt - as snowmelt got
later in the year, ignition date decreased. As DC increased, time since
snowmelt increased - this suggests that decreases in fuel moisture
significantly are a strong predictor of ignition and that ignitions were
more likely later in the year when drought code increased. SDD was
significantly influence by Biomass, age and topography; - as biomass
increased, SDD was earlier - as age increased, SDD was later - as
topography increased, sdd was earlier

Produce a clean summary table and write to csv - code from [@ramus2022]

```{r, Extreme All of Ontario Clean Summary Table}
modsum <- summary(sem_ext_2)$coefficients
modsum$Estimate <- round(modsum$Estimate, 3)
modsum$Std.Error <- round(modsum$Std.Error, 3)
modsum$Crit.Value <- round(modsum$Crit.Value, 3)
modsum$P.Value <- round(modsum$P.Value, 3)
modsum$P.Value <- ifelse(modsum$P.Value<0.001, "<0.001", modsum$P.Value)
modsum$Std.Estimate <- round(modsum$Std.Estimate, 3)
modsum$Response <- gsub("_", " ", modsum$Response)
modsum$Predictor <- gsub("_", " ", modsum$Predictor)
colnames(modsum)[4:8] <- c("SE", "DF", "t-Value", "P-Value", "Std. Estimate")
modsum

write.csv(modsum, "results/results-summary-tables/Table_S1.csv")
```

### Median burn severity for all of Ontario

```{r}
sem_med <- psem(
  lme(RBR_median ~ dc + tssm + sdd +tri +age + 
        avgBio, random = list(size_class =~1, Fire_Year =~1), na.action = na.omit, data = new_data, method = "ML"),
  lme(tssm ~ sdd + tri, random = list(size_class =~1, Fire_Year =~1), na.action = na.omit, data = new_data, method = "ML"),
  lme(sdd ~ avgBio + age + tri,random = list(size_class =~1, Fire_Year =~1), na.action = na.omit, data = new_data, method = "ML")
)


#summary
summary(sem_med)
```

Independence claim suggest significant impact of drought code on time
since snowmelt - update model to include pathway of tssm \<- dc

```{r}
sem_med_2 <- psem(
  lme(RBR_median ~ dc + tssm + sdd +tri +age + 
        avgBio, random = list(size_class =~1, Fire_Year =~1), na.action = na.omit, data = new_data, method = "ML"),
  lme(tssm ~ sdd + tri +dc + age,random = list(size_class =~1, Fire_Year =~1), na.action = na.omit, data = new_data, method = "ML"),
  lme(sdd ~ avgBio + age + tri, random = list(size_class =~1, Fire_Year =~1), na.action = na.omit, data = new_data, method = "ML")
)


#summary
summary(sem_med_2)

lapply(sem_med_2, function(i) hist(resid(i))) # looks good

```

SDD has an effect on median burn severity , as snowmelt increased median
burn severity decreased. - the difference between this and extreme is
consistent with the literature that suggests that more extreme fire
events are caused by extreme fire weather and when fire weather is
extreme it minimizes or negates the impact of other drivers. I.e.
extreme fire weather is more powerful and overcomes in the importance of
other drivers. Therefore it is expected that extreme severity events
occur under extreme fire conditions making snowmelt timing impact
negligible.

Clean summary table - code from [@ramus2022]

```{r}
modsum <- summary(sem_med_2)$coefficients
modsum$Estimate <- round(modsum$Estimate, 3)
modsum$Std.Error <- round(modsum$Std.Error, 3)
modsum$Crit.Value <- round(modsum$Crit.Value, 3)
modsum$P.Value <- round(modsum$P.Value, 3)
modsum$P.Value <- ifelse(modsum$P.Value<0.001, "<0.001", modsum$P.Value)
modsum$Std.Estimate <- round(modsum$Std.Estimate, 3)
modsum$Response <- gsub("_", " ", modsum$Response)
modsum$Predictor <- gsub("_", " ", modsum$Predictor)
colnames(modsum)[4:8] <- c("SE", "DF", "t-Value", "P-Value", "Std. Estimate")
modsum

write.csv(modsum, "results/results-summary-tables/Table_S2.csv")
```

## **What happens when we look at pyrozone patterns?**

### Western Boreal Shield of Ontario

```{r, Extreme Burn Severity Western Boreal Shield Of Ontario}

west_sem_ext <- psem(
  lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
        avgBio , random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_w, method = "ML"),
  lme(tssm ~ sdd + tri + dc, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_w, method = "ML"),
  lme(sdd ~  avgBio + age + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_w, method = "ML")
)

#summary
summary(west_sem_ext)

#examine plots
lapply(west_sem_ext, function(i) hist(resid(i))) # looks generally normal

```

### east- extreme

```{r}
east_sem_ext <- psem(
  lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
        avgBio + cc , random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_e, method = "ML"),
  lme(tssm ~ sdd + tri + dc + age, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_e, method = "ML"),
  lme(sdd ~  avgBio + age + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_e, method = "ML")
)

#cc may be better than avgBio here
#summary
summary(east_sem_ext)

###### looks like we have  an independence claim
#tssm ~ age - this is different from the west model and the full model



#examine plots
plot(east_sem_ext)
lapply(east_sem_ext, function(i) hist(resid(i))) # looks generally norma
```

### West - median

```{r, median burn severity }
west_sem_med <- psem(
  lme(RBR_median ~ dc + tssm + sdd +tri +age + 
        avgBio , random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_w, method = "ML"),
  lme(tssm ~ sdd + tri + dc + age, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_w, method = "ML"),
  lme(sdd ~  avgBio + age + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_w, method = "ML")
)

#summary
summary(west_sem_med)

#examine plots
lapply(west_sem_med, function(i) hist(resid(i))) # looks generally normal
```
### East SEM - med

```{r}
east_sem_med <- psem(
  lme(RBR_median ~ dc + tssm + sdd +tri +age + 
        avgBio , random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_e, method = "ML"),
  lme(tssm ~ sdd + tri + dc + age, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_e, method = "ML"),
  lme(sdd ~  avgBio + age + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data_e, method = "ML")
)

#summary
summary(east_sem_med)

#examine plots
lapply(east_sem_med, function(i) hist(resid(i))) # looks generally normal
```



RBR cv negative skewed
```{r}
new_data_cv <- new_data %>% mutate(rbr_cv= log10(RBR_cv))
hist(new_data_cv$rbr_cv)
```

```{r}
sem_cv <- psem(
  lme(rbr_cv ~ dc + tssm + sdd +tri +age + 
        avgBio , random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_dat_1, method = "ML"),
  lme(tssm ~ sdd + tri + dc + age +avgBio, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_dat_1, method = "ML"),
  lme(sdd ~  avgBio + age + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_dat_1, method = "ML")
)
#summary
summary(sem_cv)
```


######## Notes

When looking at the results from the updated sem, we find that: 1. DC
has a significant positive effect on extreme burn severity, we find a 2.
time since snow melt had a significant negative effect on extreme burn
severity , for every 1 day that the fire burned closer to the snow melt
date, we save decrease in burn severity. 3. sdd had a significant
negative effect on extreme burn severity, -1.8 coeff 4. age, avgBio and
cc all has significant effects, age was most significant with every 1.9
year increase in age, burn severity increased. 5. as biomass decreased
burn severity descreased, for everty 2% decrease in burn severity we saw
a decrease in burn severity 6. as CC increased burn severity increased,
for every 2 % increase in canopy cover we observed an increase in burn
severity

Our model also showed that time since snow melt is - 1. significantly
affect by snow disturbance date, for every day that snowmelt timing day
becomes earlier, we observed a decrease in tssm. suggesting that burn
day, or ignition became earlier based on snowmelt. Therefore snowmelt
might lead to earlier fires. This relationship is represented in the
literature 2. time since snowmelt was significantly effect by terrain
ruggedness index -

SDD is significantly effect by: cc *avgbio* **age**\* tri\*\*

Our hypothesized relationship structure only explained 11% (17%) of the
variance in extreme burn severity. However, SDD had a stronger
relationship with ext severity than DC.

Our hypothesized relationship structure only explained 28% (67%) of the
variance in tssm. This suggests that tssm which could be thought of as a
proxy for ignition, can be moderately explained by our relationship
structure.
