---
title: "ontario-wide-sem"
author: "Jack Goldman"
date: "2023-03-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/jgoldman/Work/PhD/forest-snow-fire-interactions/")

```

## Data analysis for structural equation Model

Load packages

```{r}
library(piecewiseSEM)
library(nlme)
library(tidyverse)
library(kableExtra)
library(ggplot2)
```

Source functions

```{r}
#indirect effects function (call ind.eff)
source("scripts/indirect-effects-function.R")

#total effects function (call tot.eff)
source("scripts/total-effects-function.R")

#vif function (call vif.lme)
source("scripts/vif-nlme-function.R")

#
source("scripts/sem-table-function.R")

```


Read in data

```{r}
df_snow <- read.csv("Data/0-forest-snow-fire-data.csv") %>% 
  filter(Fire_Year < 2020 | Fire_Year <=2001) 

#where are nas
nas <-  df_snow %>% 
  select_if(function(x) any(is.na(x))) #ecoregion na's

#omit na's
new_data <- df_snow %>% select(-c(ecoregion)) %>% na.omit(.)

#make sure DC is greater than 0
new_data <- new_data %>% filter( dc > 0) 



```

RBR quant is negatively skewed. Fix this using square root
transformation for negatively skewed data

```{r}
#sqrt transform rbr quant
new_data <- new_data %>% mutate(rbr_qs = sqrt((max(RBR_quant+1)-RBR_quant)))
hist(new_data$rbr_qs)
```


Check the distribution of SDD and TSSM

```{r}
# check distribution SDD
hist(new_data$sdd)
qqnorm(new_data$sdd)

#TSSm
hist(new_data$tssm)
qqnorm(new_data$tssm)


```



*Notes:* Fit structural equation model - all fires: - I removed climate
data because DC takes into consideration ppt and temperature from the
first day of the fire season. - Should snow Accumulation day be included
because of Chelene's past work? Chelene found that Overwintering DC has
an effect, therefore the day that snow fell would determine
overwintering DC. -DC is correlated with time since snow melt because DC
is cumulative and would differ based on the amount of days between the
start of the fire season and fire burn, should theoretically increase as
TSSM increases - TSSM includes burn day, therefore ignition
probabilities are likely influenced by canopy closure, bioamass and age.
As biomass increases, ignition probability is higher, and as canopy
closure increases, ignition and spread is more ikely and as age
increases, more flammable fuel.



Test individual model pathways for multicollinearity using VIF function

```{r, Extreme burn severity}

#VIF cutoff = 10

vif.lme(lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
    avgBio + cc, random = list(size_class =~1, Fire_Year =~1), data = new_data))

vif.lme(lme(tssm ~ sdd + dc + age + cc + avgBio , random = list(size_class =~1, Fire_Year =~1), data = new_data))


vif.lme(lme(sdd ~ cc + avgBio + age + tri , random = list(size_class =~1, Fire_Year =~1), data = new_data))



```



```{r, Median Burn Severity}
#vif cutoff = 10
vif.lme(lme(RBR_median ~ dc + tssm + sdd +tri +age + 
    avgBio + cc, random = list(size_class =~1, Fire_Year =~1), data = new_data))

vif.lme(lme(tssm ~ sdd + dc , random = list(size_class =~1, Fire_Year =~1), data = new_data))

vif.lme(lme(sdd ~ avgBio + age + cc + tri , random = list(size_class =~1, Fire_Year =~1), data = new_data))

```

Look at each models residual vs. fitted plot
```{r}
plot(lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
         avgBio + cc, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"))
plot(lme(RBR_median ~ dc + tssm + sdd +tri +age + 
         avgBio + cc, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"))
plot(lme(tssm ~ sdd + dc + age, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"))

plot(lme(sdd ~  avgBio + age + cc + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"))

```


# All of Ontario

### Extreme

```{r, Extreme all of Ontario}
sem_ext <- psem(
  lme(sdd ~  avgBio + age + cc + tri, 
      random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
  lme(tssm ~ sdd  + dc + age, 
      random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
   lme(rbr_qs ~ dc + tssm + sdd +tri +age + cc + avgBio, 
      random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML")
  
)


#summary
summary(sem_ext)

```

Original model had no effect of age on tssm. However dsep test means suggests 
that age is dseparated which means that it should be included in the model.
As age increases, snow free duration increases. 



Produce a clean summary table and write to csv - code from [@ramus2022]

```{r, Extreme All of Ontario Clean Summary Table}
modsum <- summary(sem_ext)$coefficients
modsum$Estimate <- round(modsum$Estimate, 3)
modsum$Std.Error <- round(modsum$Std.Error, 3)
modsum$Crit.Value <- round(modsum$Crit.Value, 3)
modsum$P.Value <- round(modsum$P.Value, 3)
modsum$P.Value <- ifelse(modsum$P.Value<0.001, "<0.001", modsum$P.Value)
modsum$Std.Estimate <- round(modsum$Std.Estimate, 3)
modsum$Response <- gsub("_", " ", modsum$Response)
modsum$Predictor <- gsub("_", " ", modsum$Predictor)
colnames(modsum)[4:8] <- c("SE", "DF", "t-Value", "P-Value", "Std. Estimate")
modsum <- data.frame(modsum)
modsum <- modsum %>% mutate(Response = case_when(
  Response == "rbr qs" ~"Severity",
  Response == "sdd" ~ "SFD",
  Response == "tssm" ~ "SFDr"),
  Predictor = case_when(
  Predictor == "tri" ~ "Topo",
  Predictor == "dc" ~ "Drought",
  Predictor == "cc" ~ "Canopy Closure",
  Predictor == "age" ~ "Stand Age",
  Predictor == "avgBio" ~ "Biomass", 
  Predictor == "sdd" ~ "SFD",
  Predictor == "tssm" ~ "SFDR",
  Predictor == "rbr qs" ~ "Severity"
))
colnames(modsum)[9] <-""


modsum %>% 
  kbl(caption = "piecewiseSEM results for burn severity extremes in Ontario Boreal Shield", 
      booktabs = T) %>% 
  kable_classic(full_width = F, html_font = "Cambria")

modsum <- sem_results(sem_ext)
modsum <- sem_table(sem_ext)
modsum
saveRDS(sem_ext, "results/results-summary-tables/Table-1-sem-ext-ont.rds")
```

calculate indirect effects 

```{r}
ext_ont_ind_effects <- ind.eff(sem_ext, response = "extreme")

ext_ont_ind_effects
```


Calculate total effects
```{r}
ext_ont_tot_effects <- tot.eff(sem_ext, response = "extreme")

ext_ont_tot_effects
```



### Median burn severity for all of Ontario

```{r}
sem_med <- psem(
    lme(sdd ~ avgBio + age + cc +tri,
      random = ~1 |Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
  lme(tssm ~ sdd + dc + age, 
      random = ~1 |Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
   lme(RBR_median ~ dc + tssm + sdd +tri +age + cc +avgBio, 
      random = ~1 |Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML")
)


#summary
summary(sem_med)
```


Clean summary table - code from [@ramus2022]

```{r}
modsum_med_all <- summary(sem_med)$coefficients
modsum_med_all$Estimate <- round(modsum_med_all$Estimate, 3)
modsum_med_all$Std.Error <- round(modsum_med_all$Std.Error, 3)
modsum_med_all$Crit.Value <- round(modsum_med_all$Crit.Value, 3)
modsum_med_all$P.Value <- round(modsum_med_all$P.Value, 3)
modsum_med_all$P.Value <- ifelse(modsum_med_all$P.Value<0.001, "<0.001", modsum_med_all$P.Value)
modsum_med_all$Std.Estimate <- round(modsum_med_all$Std.Estimate, 3)
modsum_med_all$Response <- gsub("_", " ", modsum_med_all$Response)
modsum_med_all$Predictor <- gsub("_", " ", modsum_med_all$Predictor)
colnames(modsum_med_all)[4:8] <- c("SE", "DF", "t-Value", "P-Value", "Std. Estimate")
modsum_med_all <- data.frame(modsum_med_all)
modsum_med_all<- modsum_med_all %>% mutate(Response = case_when(
  Response == "RBR median" ~"Severity",
  Response == "sdd" ~ "SFD",
  Response == "tssm" ~ "SFDr"),
  Predictor = case_when(
  Predictor == "tri" ~ "Topo",
  Predictor == "dc" ~ "Drought",
  Predictor == "cc" ~ "Canopy Closure",
  Predictor == "age" ~ "Stand Age",
  Predictor == "avgBio" ~ "Biomass", 
  Predictor == "sdd" ~ "SFD",
  Predictor == "tssm" ~ "SFDR",
  Predictor == "RBR median" ~ "Severity"
))
colnames(modsum_med_all)[9] <-""

modsum_med_all %>% 
  kbl(caption = "piecewiseSEM results for median burn severity in Ontario Boreal Shield", 
      booktabs = T) %>% 
  kable_classic(full_width = F, html_font = "Cambria")

modsum_med_all <- sem_table(sem_med)
modsum_med_all
saveRDS(sem_med, "results/results-summary-tables/Table-2-sem-med-ont.rds")
```

total indirect effects
```{r}

med_ont_ind_effects <- ind.eff(sem_med, response = "median")

med_ont_ind_effects 

```

specific indirect effects

```{r}
med_ont_spec_ind_eff <- spec.ind.eff(sem_med, response = "median")

med_ont_spec_ind_eff
```

total direct effects
```{r}
med_ont_tot_effects <- tot.eff(sem_med, response = "median")

med_ont_tot_effects
```
It seems that the signs flipped between total indirect effects and total effects in this equation
check to see if it makes sense

plot the indirect effects of forest structure on burn severity
```{r, median burn severity}

med_ont_eff <-  med_ont_ind_effects %>% 
  left_join(med_ont_tot_effects, by = c("Pathway", "Response")) 
med_ont_eff$`Indirect Effect` <- round(med_ont_eff$`Indirect Effect`, digits = 3)
med_ont_eff$`Total Causal Effect` <- round(med_ont_eff$`Total Causal Effect`, digits = 3)

med_ont_eff_long <- med_ont_eff %>% select(-c(Response)) %>% 
  pivot_longer(!Pathway, names_to = "type", values_to = "effect")

 
#plot
med.p.all <- ggplot()+
  geom_point(data = med_ont_eff_long, aes(x = effect, y = Pathway, colour = type), size = 5)+
theme_bw()+
  theme(axis.title = element_text(size = 14, family = "Helvetica"), 
        axis.text = element_text(size = 10, family = "Helvetica"),
        plot.title = element_text(hjust = 0.5),
    panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1)) + 
  theme(legend.position ="bottom") + xlab("Standardized Effects")+
  scale_colour_viridis_d("Effect Type", option = "viridis") +
  labs(title = "Median Burn Severity") + 
  scale_y_discrete(position = "right") #y to the right to make proper pannel of plots




# calculating confidence intervals for the random variables (path coefficients?)
med.p.all 

med.p.all <-  med.p.all + theme(legend.position = "none",
                                axis.title.y = element_blank())


```

```{r,extreme burn severity}
ext_ont_eff <-  ext_ont_ind_effects %>% 
  left_join(ext_ont_tot_effects, by = c("Pathway", "Response")) 
ext_ont_eff$`Indirect Effect` <- round(ext_ont_eff$`Indirect Effect`, digits = 3)
ext_ont_eff$`Total Causal Effect` <- round(ext_ont_eff$`Total Causal Effect`, digits = 3)

ext_ont_eff_long <- ext_ont_eff %>% select(-c(Response)) %>% 
  pivot_longer(!Pathway, names_to = "type", values_to = "effect")

 
#plot
ext.p.all <- ggplot()+
  geom_point(data = ext_ont_eff_long, aes(x = effect, y = Pathway, colour = type), size = 5)+
theme_bw()+
  theme(axis.title = element_text(size = 14, family = "Helvetica"), 
        axis.text = element_text(size = 10, family = "Helvetica"),
        plot.title = element_text(hjust = 0.5),
    panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1)) + 
  theme(legend.position ="bottom") + xlab("Standardized Effects")+
  scale_colour_viridis_d("Effect Type") +
  labs(title = "Burn Severity Extremes ") #scale_y_discrete(position = "right") #y to the right to make proper pannel of plots

ext.p.all

```

```{r}
library(patchwork)

ext.p.all + med.p.all + plot_layout(guides = "collect") & theme(legend.position = 'bottom')
ggsave("~/Desktop/OneDrive - University of Toronto/Projects/forest-snow-fire-interactions/figures-tables/figures/FigureS1-all-ind-effects.png", width = 20, height = 20, units = "cm")
```

