---
title: "ontario-wide-sem"
author: "Jack Goldman"
date: "2023-03-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/jgoldman/Work/PhD/forest-snow-fire-interactions/")

```

## Data analysis for structural equation Model

Load packages

```{r}
library(piecewiseSEM)
library(nlme)
library(tidyverse)
```

Source functions

```{r}
#indirect effects function (call ind.eff)
source("scripts/indirect-effects-function.R")

#total effects function (call tot.eff)
source("scripts/total-effects-function.R")

#vif function (call vif.lme)
source("scripts/vif-nlme-function.R")

```


Read in data

```{r}
df_snow <- read.csv("Data/0-forest-snow-fire-data.csv") %>% 
  filter(Fire_Year < 2020 | Fire_Year <=2001) 

#where are nas
nas <-  df_snow %>% 
  select_if(function(x) any(is.na(x))) #ecoregion na's

#omit na's
new_data <- df_snow %>% select(-c(ecoregion)) %>% na.omit(.)

#make sure DC is greater than 0
new_data <- new_data %>% filter( dc > 0) 



```

RBR quant is negatively skewed. Fix this using square root
transformation for negatively skewed data

```{r}
#sqrt transform rbr quant
new_data <- new_data %>% mutate(rbr_qs = sqrt((max(RBR_quant+1)-RBR_quant)))
hist(new_data$rbr_qs)
```


Check the distribution of SDD and TSSM

```{r}
# check distribution SDD
hist(new_data$sdd)
qqnorm(new_data$sdd)

#TSSm
hist(new_data$tssm)
qqnorm(new_data$tssm)


```



*Notes:* Fit structural equation model - all fires: - I removed climate
data because DC takes into consideration ppt and temperature from the
first day of the fire season. - Should snow Accumulation day be included
because of Chelene's past work? Chelene found that Overwintering DC has
an effect, therefore the day that snow fell would determine
overwintering DC. -DC is correlated with time since snow melt because DC
is cumulative and would differ based on the amount of days between the
start of the fire season and fire burn, should theoretically increase as
TSSM increases - TSSM includes burn day, therefore ignition
probabilities are likely influenced by canopy closure, bioamass and age.
As biomass increases, ignition probability is higher, and as canopy
closure increases, ignition and spread is more ikely and as age
increases, more flammable fuel.



Test individual model pathways for multicollinearity using VIF function

```{r, Extreme burn severity}

#VIF cutoff = 10

vif.lme(lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
    avgBio + cc, random = list(size_class =~1, Fire_Year =~1), data = new_data))

vif.lme(lme(tssm ~ sdd + dc + age + cc + avgBio , random = list(size_class =~1, Fire_Year =~1), data = new_data))


vif.lme(lme(sdd ~ cc + avgBio + age + tri , random = list(size_class =~1, Fire_Year =~1), data = new_data))



```



```{r, Median Burn Severity}
#vif cutoff = 10
vif.lme(lme(RBR_median ~ dc + tssm + sdd +tri +age + 
    avgBio + cc, random = list(size_class =~1, Fire_Year =~1), data = new_data))

vif.lme(lme(tssm ~ sdd + dc , random = list(size_class =~1, Fire_Year =~1), data = new_data))

vif.lme(lme(sdd ~ avgBio + age + cc + tri , random = list(size_class =~1, Fire_Year =~1), data = new_data))

```

Look at each models residual vs. fitted plot
```{r}
plot(lme(rbr_qs ~ dc + tssm + sdd +tri +age + 
         avgBio + cc, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"))
plot(lme(tssm ~ sdd + dc + age, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"))

plot(lme(sdd ~  avgBio + age + cc + tri, random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"))

```


# All of Ontario

### Extreme

```{r, Extreme all of Ontario}
sem_ext <- psem(
  lme(rbr_qs ~ dc + tssm + sdd +tri +age + cc + avgBio, 
      random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
  lme(tssm ~ sdd  + dc + age, 
      random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML"),
  lme(sdd ~  avgBio + age + cc + tri, 
      random = ~1 | Fire_Year/size_class, na.action = na.omit, data = new_data, method = "ML")
)


#summary
summary(sem_ext)

```

Produce a clean summary table and write to csv - code from [@ramus2022]

```{r, Extreme All of Ontario Clean Summary Table}
modsum <- summary(sem_ext)$coefficients
modsum$Estimate <- round(modsum$Estimate, 3)
modsum$Std.Error <- round(modsum$Std.Error, 3)
modsum$Crit.Value <- round(modsum$Crit.Value, 3)
modsum$P.Value <- round(modsum$P.Value, 3)
modsum$P.Value <- ifelse(modsum$P.Value<0.001, "<0.001", modsum$P.Value)
modsum$Std.Estimate <- round(modsum$Std.Estimate, 3)
modsum$Response <- gsub("_", " ", modsum$Response)
modsum$Predictor <- gsub("_", " ", modsum$Predictor)
colnames(modsum)[4:8] <- c("SE", "DF", "t-Value", "P-Value", "Std. Estimate")
modsum

write.csv(modsum, "results/results-summary-tables/Table-1-sem-ext-ont.csv")
```

calculate indirect effects 

```{r}
ext_ont_ind_effects <- ind.eff(modsum, response = "extreme")

ext_ont_ind_effects
```
Specific indirect effects
```{r}
ext_spec_ind_eff <- spec.ind.eff(modsum, response = "extreme")

ext_spec_ind_eff
```


Calculate total effects
```{r}
ext_ont_total_effects <- tot.eff(modsum, response = "extreme")

ext_ont_total_effects
```



### Median burn severity for all of Ontario

```{r}
sem_med <- psem(
  lme(RBR_median ~ dc + tssm + sdd +tri +age + cc +avgBio, 
      random = list(size_class =~1, Fire_Year =~1), na.action = na.omit, data = new_data, method = "ML"),
  lme(tssm ~ sdd + dc + age,
      random = list(size_class =~1, Fire_Year =~1), na.action = na.omit, data = new_data, method = "ML"),
  lme(sdd ~ avgBio + age + cc +tri,
      random = list(size_class =~1, Fire_Year =~1), na.action = na.omit, data = new_data, method = "ML")
)


#summary
summary(sem_med)
```


Clean summary table - code from [@ramus2022]

```{r}
modsum_med_all <- summary(sem_med)$coefficients
modsum_med_all$Estimate <- round(modsum_med_all$Estimate, 3)
modsum_med_all$Std.Error <- round(modsum_med_all$Std.Error, 3)
modsum_med_all$Crit.Value <- round(modsum_med_all$Crit.Value, 3)
modsum_med_all$P.Value <- round(modsum_med_all$P.Value, 3)
modsum_med_all$P.Value <- ifelse(modsum_med_all$P.Value<0.001, "<0.001", modsum_med_all$P.Value)
modsum_med_all$Std.Estimate <- round(modsum_med_all$Std.Estimate, 3)
modsum_med_all$Response <- gsub("_", " ", modsum_med_all$Response)
modsum_med_all$Predictor <- gsub("_", " ", modsum_med_all$Predictor)
colnames(modsum_med_all)[4:8] <- c("SE", "DF", "t-Value", "P-Value", "Std. Estimate")
modsum_med_all

write.csv(modsum_med_all, "results/results-summary-tables/Table-2-sem-med-ont.csv")
```

total indirect effects
```{r}

med_ont_ind_effects <- ind.eff(modsum_med_all, response = "median")

med_ont_ind_effects 

```

specific indirect effects

```{r}
med_ont_spec_ind_eff <- spec.ind.eff(modsum_med_all, response = "median")

med_ont_spec_ind_eff
```

total direct effects
```{r}
med_ont_tot_effects <- tot.eff(modsum_med_all, response = "median")

med_ont_tot_effects
```
It seems that the signs flipped between total indirect effects and total effects in this equation
check to see if it makes sense

plot the indirect effects of forest structure on burn severity
```{r}
med_ont_ind_eff_sf <- med_ont_ind_effects %>% 
  filter(Mediator == "snow free date") %>% 
  mutate(`Snow Free Date` = rep("Snow Free Date"))
med_ont_ind_eff_sfd <- med_ont_ind_effects %>% 
  filter( Mediator == "snow free duration")%>% 
  mutate(`Snow Free Duration` = rep("Snow Free Duration"))
med_ont_tot_effects <- med_ont_tot_effects %>% 
  rename(tot_ind_eff = "Total Indirect Effect") %>% 
  rename(tot_eff ="Total Effect") %>% 
  mutate(`Total Indirect Effect` = rep("Total Indirect Effect"),
                                     `Total Effect`= rep("Total Effect"))

 ggplot()+
  geom_point(data = med_ont_ind_eff_sf, aes(x = `Indirect Effect`, y = Pathway, color = "Snow Free Date"), size = 5) +
  geom_point(data = med_ont_ind_eff_sfd, aes(x = `Indirect Effect`, y = Pathway, color = "Snow Free Duration"), size = 5) +
  geom_point(data = med_ont_tot_effects, aes(x = tot_ind_eff, y = Pathway, color = "Total Indirect Effect"), size = 5)+ 
  geom_point(data = med_ont_tot_effects, aes(x= tot_eff, y = Pathway, color = "Total Effect"), size =5) + 
  theme_bw()+
  theme(axis.title = element_text(size = 14, family = "Helvetica"), 
        axis.text = element_text(size = 10, family = "Helvetica"),
    panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1)) + 
  theme(legend.position = c(0.8, 0.87))




```

